---
title: "LOGISTIC REGRESSION Problem Set #2"
author: "Patrick Torralba"
format: html
editor: visual
---

# Logistic Regression Notes

## Which of the assumptions are checked for Logistic Regression?

-   Linearity

-   Independent observations

-   Normal distribution

-   Equal variance

-   basically LINE but on the log scale

## What is the null hypothesis for a coefficient in a logistic regression model...

On the log-odds scale? H0: beta = 0; beta is log(odds)

On the odds ratio scale? H0: exp(beta) = 1; exp(beta) = odds ratio

## What are two ways we can determine statistical significance of a coefficient in a logistic regression model?

1.  Confidence Interval does not include 1
2.  There is a dramatic change in the parameter coefficient

# Homework

```{r load library, include=FALSE, echo=FALSE}
library(epiDisplay)
library(tidyverse)
library(broom)
library(writexl)
```

```{r import data, include=FALSE, echo=FALSE}
url <- "https://raw.githubusercontent.com/torra034-svg/biostats/refs/heads/main/Biostats%202/Data%20Stuff/dekaf_dgf.csv"

dekaf <- read.csv(url)
```

```{r remove url, echo=FALSE}

rm(url)

```

# Exploratory Data Analysis

```{r Table 1}

table_1 <- dekaf %>% 
  dplyr::select(pid, dgf, wit_unidist, daysondial, wtkg_donor, bmi_base) 

table_1
```

# Odds and Odds Ratios

```{r 2x2 table - use epiDisplay (more easy)}

matrix_table <- cc(dekaf$dgf, dekaf$gender, graph = FALSE)


matrix_table
```

note: odds means that we are comparing one portion to another portion. Not the total. AKA you divide exposed / unexposed or however the question asks you.

other note: The odds ratio is taking the (odds of group 1) / (odds of group 2) \<-- depends on the question

# MODEL 1: Logistic model for DGF by recipient gender

```{r Logistic Regression Model by Gender}

log_model <- glm(data = dekaf, dgf ~ gender, family = "binomial")


summary(log_model)
```

log model note: To run a logistic regression in R, use the glm() function and make sure to specify the family as binomial.

-   When writing down fitted model, you cannot simply write log(y) = intercept + (B1 x X1)
-   You must write logit(\[P(Y = 1)\]) = intercept + (B1 x X1)
-   logit means log of the odds
-   You must write it in this form to since the we cannot simply take the log of a binary outcome, we must specify that the binary outcome is = 1 (or however it is coded)
-   Make sure when interpreting the coefficient, you must convert the log coefficient back to normal back exponentiating it (e.g j raise to the power of e)
-   Same thing when interpreting the odds ratio from the fitted formula. ALWAYS convert back to base by expatiating the values for interpretation.

# MODEL 2: Logistic model for DGF by days on dialysis prior to transplant

```{r logistic regresion model - DGF by days on dialysis}

log_model_2 <- glm(data = dekaf, dgf ~ daysondial, family = "binomial")


summary(log_model_2)
```

```{r Confidence Interval of Odds Ratio}

exp(confint(log_model_2))

```

-   Make sure to exponentiate the confidence interval to convert from log to normal values.
-   Hint: An Odds Ratio can never be negative. Thus, the confidence interval can only range from 0 to infinity in which 1 = no effect.

# Model 3: Logistic model for DGF by days on dialysis adjusting for transfusions prior to transplant

```{r data cleaning, echo=FALSE}

dekaf <- dekaf %>% 
  mutate(transfusions = if_else(transfusions_yn == 1, "Transfusion", "No Transfusion", missing = NULL))

```

```{r logistic regression model - adjusted for transfusions prior to transplant}

log_model_2_adjusted <- glm(data = dekaf, dgf ~ daysondial*transfusions, family = "binomial")

summary(log_model_2_adjusted)
```

-   Note: since presence of transfusions may be a possible confounder, we must account for this interaction term.
-   Other note about glm and log: **ONLY EXPONENTIATE THE COEFFICIENTS**
-   Also make sure to set **family = "binomial"** \<-- this step is super important for glm
-   To check if a variable is a confounder, make sure to look at both the coefficient change AND the p-value.

```{r Use tidy function from broom package to make things tidy!}

# tidy the adjusted log model
tidy_log_model_2_adjusted <- tidy(log_model_2_adjusted, conf.int = TRUE, exponentiate = TRUE)

tidy_log_model_2_adjusted
```

```{r export tidy adjusted log model into a clean excel file for hw purposes, include=FALSE, echo=FALSE}

write_xlsx(tidy_log_model_2_adjusted, "DGF_adjusted_model.xlsx")

```

-   Note: I like the tidy function! It makes my output table look more neat and readable. I also don't need to manually exponentiate each coefficient estimate.
-   Need to keep the summary() log version to write the logit fitted model.

## What are the estimated odds of DGF for someone with a transfusion and with 30 days of dialysis prior to transplant?

```{r q30 - use predict function and new data frame }

q30 <- data.frame(daysondial = 30, transfusions = "Transfusion")

q30_log_odds <- exp(predict(log_model_2_adjusted, newdata = q30, type = "link"))

q30_log_odds
```

-   for this question, use the **original log model** since it's easier to input values.
-   Do not forget to exponentiate values

## Using Model 3, what are the estimated odds of DGF for someone without a transfusion and with 30 days of dialysis prior to transplant?

```{r q31 - use predict function again}

q31 <- data.frame(daysondial = 30, transfusions = "No Transfusion")

q31_log_odds <- exp(predict(log_model_2_adjusted, newdata = q31, type = "link"))

q31_log_odds
```

Calculate the odds ratio comparing the odds of DGF for someone with a transfusion to someone without a transfusion, both having had 30 days of dialysis prior to transplant

```{r odds ratio transfusion}

odds_ratio_transfusion <- (q30_log_odds / q31_log_odds)

odds_ratio_transfusion
```

Provide an estimate of the adjusted odds ratio comparing those with a transfusion compared to those without, adjusting for days on dialysis, together with its 95% confidence interval.

```{r adjusted odds ratio estimate calculation}

log_model_adj_final <- glm(data = dekaf, 
                           dgf ~ transfusions*daysondial,
                           family = "binomial")

log_model_adj_final
```

```{r tidy adjusted odds-ratio model}

tidy_final <- tidy(log_model_adj_final,
                   conf.int = TRUE,
                   exponentiate = TRUE)


tidy_final
```

Provide an estimate of the adjusted odds ratio comparing the contribution of an additional day on dialysis, adjusted for the presence of transfusions, and its 95% confidence interval.
```{r adjusted odds-ratio daysondial*transfusions}

log_model_adj_day_trans <- glm(
  data = dekaf,
  dgf ~ daysondial*transfusions,
  family = "binomial"
)

log_model_adj_day_trans

```

```{r tidy log_model_adj_day_trans}

tidy_adj_model_day_trans <- tidy(log_model_adj_day_trans,
                                 conf.int = TRUE,
                                 exponentiate = TRUE)


tidy_adj_model_day_trans
```

