---
title: "LOGISTIC REGRESSION Problem Set #2"
author: "Patrick Torralba"
format: html
editor: visual
---

```{r load library, include=FALSE, echo=FALSE}
library(epiDisplay)
library(tidyverse)
```

```{r import data, include=FALSE, echo=FALSE}
url <- "https://raw.githubusercontent.com/torra034-svg/biostats/refs/heads/main/Biostats%202/Data%20Stuff/dekaf_dgf.csv"

dekaf <- read.csv(url)
```

```{r remove url, echo=FALSE}

rm(url)

```

#Exploratory Data Analysis

```{r Table 1}

table_1 <- dekaf %>% 
  dplyr::select(pid, dgf, wit_unidist, daysondial, wtkg_donor, bmi_base) 

table_1
```

# Odds and Odds Ratios

```{r 2x2 table - use epiDisplay (more easy)}

matrix_table <- cc(dekaf$dgf, dekaf$gender, graph = FALSE)


matrix_table
```

note: odds means that we are comparing one portion to another portion. Not the total. AKA you divide exposed / unexposed or however the question asks you.

other note: The odds ratio is taking the (odds of group 1) / (odds of group 2) \<-- depends on the question

# MODEL 1: Logistic model for DGF by recipient gender

```{r Logistic Regression Model by Gender}

log_model <- glm(data = dekaf, dgf ~ gender, family = "binomial")


summary(log_model)
```

log model note: To run a logistic regression in R, use the glm() function and make sure to specify the family as binomial.

-   When writing down fitted model, you cannot simply write log(y) = intercept + (B1 x X1)
-   You must write logit(\[P(Y = 1)\]) = intercept + (B1 x X1)
-   logit means log of the odds
-   You must write it in this form to since the we cannot simply take the log of a binary outcome, we must specify that the binary outcome is = 1 (or however it is coded)
-   Make sure when interpreting the coefficient, you must convert the log coefficient back to normal back exponentiating it (e.g j raise to the power of e)
-   Same thing when interpreting the odds ratio from the fitted formula. ALWAYS convert back to base by expatiating the values for interpretation.

# MODEL 2: Logistic model for DGF by days on dialysis prior to transplant

```{r logistic regresion model - DGF by days on dialysis}

log_model_2 <- glm(data = dekaf, dgf ~ daysondial, family = "binomial")


summary(log_model_2)
```


```{r Confidence Interval of Odds Ratio}

exp(confint(log_model_2))

```

* Make sure to exponentiate the confidence interval to convert from logg to normal values.
* Hint: An Odds Ratio can never be negative. Thus, the confidence interval can only range from 0 to infinity in which 1 = no effect.


# Model 3: Logistic model for DGF by days on dialysis adjusting for transfusions prior to transplant

```{r data cleaning, echo=FALSE}

dekaf <- dekaf %>% 
  mutate(transfusions = if_else(transfusions_yn == 1, "Transfusion", "No Transfusion", missing = NULL))

```


```{r logistic regression model - adjusted for transfusions prior to transplant}

log_model_2_adjusted <- glm(data = dekaf, dgf ~ daysondial*transfusions, family = "binomial")

summary(log_model_2_adjusted)
```

* Note: since presence of transfusions may be a possible confounder, we must account for this interaction term.
* **Other note about glm and log: ONLY EXPONENTIATE THE COEFFICIENTS**
* Also make sure to set **family = "binomial"** <-- this step is super important for glm