---
title: "Project 2: Coding Reference Sheet"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    code-fold: FALSE
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 4
    smooth-scroll: TRUE
    embed-resources: TRUE
    number-sections: TRUE
editor: visual
---

# Exploratory Data Analysis Concepts

## Importing Data

### How do I import data using the tidyverse?

#### Library

To first import data into R, we need to load the tidyverse package. This package gives us the tools needed to both import and clean data.

In the code chunk below, we load in the tidyverse package.
```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true

library(tidyverse)  
  
```


#### Import Data

We use the read.csv() function to import .csv files. Make sure to specify the file pathway and assign it to a variable using "<-".

In the code chunk below, we import the tomhs dataset by specifying it's location and file extension.
```{r}
#| label: import-data
#| warning: false
#| message: false
#| echo: true

tomhs <- read.csv("Other Regression Models/Ordinal Regression/tomhs (2).csv")  


```


## Data Wrangling

### Creating New Variables

To create new variables, we first want to overwrite our dataset with changes with the "<-" operator.

Use the pipe operator to combine functions together in one code chunk.

Use mutate() to define our new variables.

Note: Use the if_else() function when you have a singular class, and use case_when() when the categorical variable has multiple classes.

In the code chunk below, we create two new variables, "Sex" and "Race". Sex is created by an if_else() statement and Race is created by a case_when() statement. Both variables are converted into factors - this is important for statistical analysis purposes.
```{r}
#| label: new-var
#| warning: false
#| message: false
#| echo: true

tomhs <- tomhs %>% 
  mutate(
    Sex = as.factor(if_else(sex == 1, "Male", "Female")),
    Race = as.factor(case_when(
      race == 1 ~ "White",
      race == 2 ~ "Black",
      race == 3 ~ "Asian",
      race == 4 ~ "Other"
    )))
  
```


### Deleting & Subsetting Observations

Deleting observations is done through the filter() function. Use this function to specify the delete observations.

General Tips
* "&" is the AND operator
* "|" is the OR operator
* "!" is the NOT operator
* "is.na()" is used to specify NA values.

We should create a new dataset just in case we want to refer to the original dataset.

In the code chunk below, we remove observations in the age column that are equivalent to NA. And, we also subset these changes into a new dataset, "tomhs_filter" to ensure the original dataset maintains the original observation amount.
```{r}
#| label: delete-obs
#| warning: false
#| message: false
#| echo: true

tomhs_filter <- tomhs %>% 
  filter(!is.na(age))
  

```


# Regression Modeling

## Multiple Linear Regression (MLR)

MLR is when we try to predict a **continuous outcome** with either a continuous predictor or categorical predictor. 

### Unadjusted MLR

### Adjusted MLR

### Interaction Effects

## Logistic Regression

## Cox Proportional Hazards Model

# Other Regression Models

## Poisson Regression Model

## Log-binomial Regression Model

## Ordinal Regression Model

# Sample Size Estimates and Power

## Continous Outcomes

## Binary Outcomes

# .qmd Notes