---
title: "Sample Size and Power"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    code-fold: FALSE
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 3
    smooth-scroll: TRUE
    embed-resources: TRUE
    number-sections: TRUE
editor: visual
---

# Set-up

## Library

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true

library(tidyverse)
library(pwr)
library(gtsummary) ## translate output to table vis tbl_regression()
library(kableExtra) ## transalte output to table via kable()
library(broom) ## translate output (basically a better version of summary)

```

# Situation 1: Cusick et al. article

## Estimating Sample Sizes in Group Means

```{r}
#| label: tbl-sample-size-gm
#| tbl-cap: "Estimating Sample Sizes from the Cusick Article"
#| tbl-cap-location: top


## pwr.t.test()
### haha jk we did not do Part 2 from section 1. Therefore, we cannot do this problem lol.


```

# Situation 2: Continuous Outcome

## Estimating Sample Sizes in Group Means {#sec-group-mean-sample-size-standard}

```{r}
#| label: tbl-sample-size-q6
#| tbl-cap: "Estimating Sample Size from a Continuous Outcome"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


power.t.test(
  delta = 2, ## this is the MCID value
  sd = 6,
  sig.level = 0.05,
  power = 0.8,
  type = "two.sample",
) %>% 
  tidy() %>% 
  kable(
        format = "html",
        digits = 2,
        align = "c") %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  )

```

## Estimating Multiple Sample Size via Power Table {#sec-power-table-sample-size}

```{r}
#| label: tbl-sample-size-q9
#| tbl-cap: "Estimating Multiple Sample Sizes via a Power Table"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true

deltas <- seq(
  from = 1,
  to = 5,
  by = 1
)

q9 <- map_df(deltas, ~{
  power.t.test(
    delta = .x,         
    sd = 6,              
    sig.level = 0.05, 
    power = 0.80, 
    type = "two.sample"
  ) %>% tidy()
})

q9 %>% 
  kable(
        format = "html",
        digits = 2,
        align = "c") %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  )
  

## DO NOT TRUST THE SOFTWARE LESSON. IT IS INACCURATE
```

## Using a Power Curve to Compare Two Different MCIDS (Effect Sizes) {#sec-power-curve-plot-two-different-MCIDS}

```{r}
#| label: fig-power-curve
#| fig-cap: "Using a Power Curve to Compare Two Different MCIDS (Effect Sizes)"
#| fig-cap-location: top
#| warning: false
#| message: false
#| echo: true

## We need to create a list of values for our x-axis in the power curve plot
n.values <- seq(from = 5,
                to = 100,
                by = 5) 

## Get Sample Size when MCID = 2
twogrpmean.power.small <-
  power.t.test(
    delta = 2,
    sd = 6,
    power = NULL, ## Leave Power NULL when calculating power curve (bc this is what we want R to graph for us)
    n = n.values,
    type = "two.sample"
  )
  

## Get Sample Size when MCID = 4
twogrpmean.power.medium <-
    power.t.test(
    delta = 4,
    sd = 6,
    power = NULL,
    n = n.values,
    type = "two.sample",
  )

## Plotting the first curve (MCID = 2)
plot(n.values,
     twogrpmean.power.small$power,
     type = "b",
     xlab = "sample size",
     ylab = "power",
     ylim = c(0,1))

lines(n.values, 
      twogrpmean.power.small$power, 
      type = "b") 

## cannot add type = "b" argument. This will cause graph not to run.


## Add Second Curve (MCID = 4)
points(n.values,
       twogrpmean.power.medium$power,
       type = "b",
       col = "red",) ## Keep these arguments like this to run. Otherwise, if you follow the software lesson it will not run.


## Add a legend for each curve
legend("bottomright", c( "d = 2", "d = 4"), lty = c(1,2),
pch = c(1,2), col = c(1,2), cex=1.5)

```

# Situation 3: Binary Outcome

## Estimating Sample Sizes {#sec-binary-outcome-sample-size-calculation}

```{r}
#| label: tbl-sample-size-group-prp
#| tbl-cap: "Estimating Sample Size in a Group Proportion"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true

twogrpp.n <- power.prop.test(
  p1 = 0.70,
  p2 = 0.80,
  power = 0.80
)

twogrpp.n %>% 
  tidy() %>% 
  kable(
        format = "html",
        digits = 2,
        align = "c") %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  )


```

## Calculating Multiple Sample Sizes with the same MCID {#sec-binary-outcome-multiple-sample-size-same-MCID}

```{r}
#| label: tbl-calc-mult-samp-size-grp-prp
#| tbl-cap: "Calculating Multiple Sample Sizes in a Group Proportion with the same MCID"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


## Define Baseline Proportions
p1_levels <- seq(from = 0.50, to = 0.75, by = 0.05)

# Level 1: 50% to 60%
res1 <- power.prop.test(p1=0.50, p2=0.60, sig.level=0.05, power=0.80, n=NULL)
n1   <- ceiling(res1$n)

# Level 2: 55% to 65%
res2 <- power.prop.test(p1=0.55, p2=0.65, sig.level=0.05, power=0.80, n=NULL)
n2   <- ceiling(res2$n)

# Level 3: 60% to 70%
res3 <- power.prop.test(p1=0.60, p2=0.70, sig.level=0.05, power=0.80, n=NULL)
n3   <- ceiling(res3$n)

# Level 4: 65% to 75%
res4 <- power.prop.test(p1=0.65, p2=0.75, sig.level=0.05, power=0.80, n=NULL)
n4   <- ceiling(res4$n)

# Level 5: 70% to 80% (Original target)
res5 <- power.prop.test(p1=0.70, p2=0.80, sig.level=0.05, power=0.80, n=NULL)
n5   <- ceiling(res5$n)

# Level 6: 75% to 85%
res6 <- power.prop.test(p1=0.75, p2=0.85, sig.level=0.05, power=0.80, n=NULL)
n6   <- ceiling(res6$n)

## 3. Combine results into the final table
twogrpp.pwrtable <- data.frame(
  Baseline_Suppression = p1_levels,
  Comparison_Suppression = p1_levels + 0.10,
  Required_N_per_Group = c(n1, n2, n3, n4, n5, n6)
)

# Display the final table
twogrpp.pwrtable


twogrpp.pwrtable %>% 
    kable(
        format = "html",
        digits = 2,
        align = "c") %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  )
```

## Using a Power Curve to Estimate Power based on Sample Size {#sec-binary-outcome-power-curve}

```{r}
#| label: fig-power-curve-q19
#| fig-cap: "Power Curve Comparing Baseline Viral Suppression Levels (60% vs 70%)"
#| fig-cap-location: top
#| warning: false
#| message: false
#| echo: true

## Define range of samples

n_range <- seq(from = 10, to = 300, by = 10)


## Get Sample Size for each proportion

pow_60 <- 
  power.prop.test(
    p1 = 0.60,
    p2 = 0.70,
    n = n_range,
  )$power

pow_70 <-
  power.prop.test(
    p1 = 0.70,
    p2 = 0.80,
    n = n_range,
  )$power

## Plot
plot(n_range, pow_70, 
     type = "b", 
     col = "blue", 
     pch = 16,
     xlab = "Sample Size (per group)", 
     ylab = "Power", 
     ylim = c(0, 1))

# Add the 60% line
lines(n_range, pow_60, type = "b", col = "red", pch = 16)

# Add a reference line for 80% power
abline(h = 0.80, lty = 2, col = "darkgrey")

## 5. Add a legend that matches your data
legend("bottomright", 
       legend = c("Baseline 70% (Target 80%)", "Baseline 60% (Target 70%)"), 
       col = c("blue", "red"), 
       pch = 16, 
       lty = 1, 
       cex = 0.8) # Adjusted cex so it fits better
```

## Calculating Sample Size Lost to Attrition {#sec-sample-size-attrition}

-   Question: You and your collaborator decide to be a bit conservative and use a viral suppression rate of 60%. You also believe based on past experience that about 20% of the enrolled participants will be lost to follow-up over the one year of the study. How many participants should you aim to enroll for your study? Please show your work.

-   Know

    -   n = 356
        -   Since we use the 60% prop, the n is 356 according to the Power Table
    -   attrition rate = 0.20
        -   This is stated in the problem \[the people lost to follow-up\]
    -   Formula to calculate **Adjusted Sample Size PER GROUP**
        -   $$
            N_{adj} = \frac{n}{1 - \text{attrition rate}}
            $$

    ```{r}
    #| label: attrition
    #| warning: false
    #| message: false
    #| echo: true

    (356/(1-0.2))


    ```




# Power and Sample Size Summary Notes

-   What if my outcome is Continuous (Group Means)?
    -   Information needed to calculate sample size:
        -   Mean
        -   Standard Deviation
        -   Effect Size (also known as delta or MCID)
        -   Power Level (usually 0.80 unless stated otherwise)
    -   How to calculate sample size?
        -   @sec-group-mean-sample-size-standard
    -   How to calculate multiple sample sizes via Power Table?
        -   @sec-power-curve-plot-two-different-MCIDS
    -   How to generate a Power Curve to compare two different MCIDS?
        -   @sec-power-curve-plot-two-different-MCIDS
-   What if my outcome is Binary (Proportions)
    -   Information needed to calculate sample size:
        -   Standard Deviations **IS NOT NEEDED**
        -   Baseline Proportion
        -   Comparison Proportion
        -   Effect Size (MCID)
    -   How to calculate sample size?
        -   @sec-binary-outcome-sample-size-calculation
    -   How to calculate multiple sample sizes via Power Table?
        -   @sec-binary-outcome-power-curve
    -   How to generate a Power Curve to Compare Two Different Proportions?
        -   @sec-binary-outcome-power-curve
-   How to calculate the adjusted sample size when accounting for people lost to follow-up?
    -   @sec-sample-size-attrition
-   
