---
title: "Project 2 - Coding Reference Sheet | Submission Version"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    fontsize: 10pt
    code-fold: FALSE
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 5
    smooth-scroll: TRUE
    embed-resources: TRUE
    fig-align: center
execute:
  results: hide
  output: false
editor: visual
---

# Exploratory Data Analysis Concepts

## Importing Data

### How do I import data using the tidyverse?

#### Library {#sec-library}

To first import data into R, we need to load the tidyverse package. This package gives us the tools needed to both import and clean data. In the code chunk below, we load in the tidyverse package (and other packages that are more relevant in other sections).

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true

library(tidyverse) ## data cleaning 
library(kableExtra) ## make table from output via kable()
library(gtsummary) ## make table via tbl_regression()
```

#### Import Data

We use the read.csv() function to import .csv files. Make sure to specify the file pathway and assign it to a variable using "\<-". In the code chunk below, we import the data by specifying it's location and file extension.

```{r}
#| label: import-data
#| warning: false
#| message: false
#| echo: true

#data <- read.csv("data.csv")  
```

## Data Wrangling

### Creating New Variables {#sec-new-var}

To create new variables, we first want to overwrite our dataset with changes with the "\<-" operator. Use the pipe operator to combine functions together in one code chunk. Use mutate() to define our new variables. In the code chunk below, we create two new variables, "Var1" and "Var2". Var1 is created by an if_else() statement and Var2 is created by a case_when() statement. Both variables are converted into factors - this is important for statistical analysis purposes.

```{r}
#| label: new-var
#| warning: false
#| message: false
#| echo: true

#data <- data %>% 
  #mutate(
   # Var1 = as.factor(if_else(var1 == 1, "A", "B")),
  #  Var2 = as.factor(case_when(var2 == 1 ~ "W",race == 2 ~ "B",race == 3 ~ "A", race == 4 ~ "O")))
    

  
```

### Deleting & Sub-setting Observations

Deleting observations is done through the filter() function. Use this function to specify which observations to delete.

| Operator |   Definition    |
|:--------:|:---------------:|
|    &     |  AND operator   |
|    \|    |   OR operator   |
|    !     |  NOT operator   |
|   \<-    | Assign variable |

: Operators

| BaseR Functions |                Definition                 |
|:---------------:|:-----------------------------------------:|
|     is.na()     | Filters to rows where there are NA values |

: BaseR Functions

We should create a new dataset just in case we want to refer to the original dataset. In the code chunk below, we remove observations that are equivalent to NA. And, we also subset these changes into a new dataset, "data_filter" to ensure the original dataset maintains the original observation amount.

```{r}
#| label: delete-obs-sub-set
#| warning: false
#| message: false
#| echo: true

#data_sub <- data %>% 
#  filter(!is.na(var1))
  
```

# Basic Exploratory Data Analysis (EDA)

EDA is an approach to analyzing and summarizing data sets with an intent to understand their main characteristics. This approach includes both tables and graphs.

## Example

### Summary Statistics

For Table 1 purposes, we only want to summarize the main characteristics of the study demographic. Do not include any significant predictor variables. We only want to include general characteristics - these are usually race, sex, BMI, etc.

```{r}
#| label: dekaf-import
#| message: false
#| warning: false
#| echo: true
#| include: false

dekaf <- read.csv("dekaf_dgf.csv")

```

```{r}
#| label: eda-example-dekaf
#| message: false
#| warning: false
#| echo: true  
#| include: false

dekaf <- dekaf %>% 
  mutate(cardiac_death = if_else(dcd_yn == 1, "Yes", "No", missing = NULL),
         DGF = if_else(dgf == 1, "Yes", "No", missing = NULL))

```

```{r}
#| label: table-1-data-clean
#| message: false
#| warning: false
#| echo: true  
#| include: false


table_1 <- dekaf %>% 
  mutate(Race = if_else(raceblack == 1, "African American", "Other", missing = NULL),
         Gender = if_else(gender == 1, "Male", "Female", missing = NULL),
         Delayed_Graft_Function = if_else(dgf == 1, "Yes", "No", missing = NULL)) %>% 
  dplyr::select(Race, Gender, Delayed_Graft_Function, cardiac_death, bmi_base, agedonor16)

```

```{r}
#| label: tbl-table1
#| tbl-cap: "Table 1: Study Demographics"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true

#table_1 <- data %>% 
  #tbl_summary(by = group_var)
   #add_overall()

#table_1
```

### Graphical Summaries

We want to utilize visualizations that summarize the relationship between the predictor variables and outcome variable so we can have some idea of what this relationship may be.

```{r}
#| label: fig-graph-summary-dekaf
#| fig-cap-location: top
#| message: false
#| warning: false
#| echo: true

#ggplot(data, aes(x = var1, y = var2, fill = var2)) +
  #geom_boxplot() 
```

# Regression Modeling

## Multiple Linear Regression (MLR) {#sec-hyp-test}

MLR is when we try to predict a **continuous outcome** with either a continuous predictor and/or categorical predictor. The null hypothesis is that all regression coefficients are equal to zero. The alternative hypothesis is that atleast one regression coefficient is not equivalent to zero.

$$
H_0: \beta_1 = \beta_2 
$$

$$
H_1: \beta_1 \neq \beta_2
$$

Unadjusted means that we do not account or adjust for other variables that may affect the outcome variable. Adjusted means we do account for other variables that affect the outcome variable.

### Unadjusted MLR {#sec-regression-model-tips}

Example Question: What is the effect of weight change on total cholesterol change without adjusting for anything In the code chunk below, we conduct an MLR by using the lm() function.

```{r}
#| label: tbl-unadj-MLR
#| tbl-cap: "Unadjusted Multiple Linear Regression"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true

#unadj_MLR <- lm(outcome_var ~ predictor_var, data = data)

#summary(unadj_MLR)
```

#### Multiple Linear Regression Fitted Equation {#sec-MLR-equation}

$$
\text{outcome-variable} = \text{intercept} + (\beta_1 X_1) + (\beta_2 X_2) + ... 
$$

#### Interpretation {#sec-MLR-interpretation}

The intercept means when the x-variable is at zero, we expect the outcome variable to be at whatever the intercept is. If the x-variable is a **continuous predictor**, for every 1 unit increase in the x-variable, we expect the y-variable to increase by the x-variable coefficient $\beta_1$. If the x-variable is a **categorical predictor** (e.g binary), and the x-variable is equivalent to 1, then we expect the y-variable to increase by the x-variable coefficient $\beta_1$. If the p-value is less than 0.05, it is statistically significant.

### Adjusted MLR {#sec-adjusted-MLR}

```{r}
#| label: tbl-Adj-MLR
#| tbl-cap: "Adjusted Multiple Linear Regression"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true
#| include: false

#adj_MLR <- lm(outcome_var ~ predictor_var + predictor_var2, data = data)

#summary(adj_MLR)
```

#### Interpretation {#sec-adj-MLR-interpretation}

This fitted equation represents the relationship between outcome variable when we both account for both predictor variables.

### Interaction Plot

Interaction plots are used to visualize the effect of one predictor variable on the outcome variable while depending on the value of another predictor variable. Essentially, they tell us if two variables are synergistic or interference.

**Question:** Make an interaction plot.

```{r}
#| label: tomhs-data-clean
#| message: false
#| warning: false
#| echo: true
#| include: false

tomhs <- read.csv("tomhs.csv")


tomhs <- tomhs %>% 
  mutate(wtchg = wt12 - wtbl,
         cholchg = chol12 - cholbl,
         nowsmk1 = as.factor(if_else(nowsmk == 1, "Current Smoker", "Non-current Smoker")))

```

```{r}
#| label: fig-interact-plot
#| fig-cap-location: top
#| message: false
#| warning: false
#| echo: true

#library(interactions)
#model <- lm(outcome_var ~ predictor_var*predictor_var2, data = data)
#interact_plot(model, pred = wtchg, modx = nowsmk1,
#              interval = TRUE,
#               plot.points = TRUE,
#               x.label = "predictor_var",
#               y.label = "outcome_var",
#               legend.main = "predictor_var2",           
#               main.title = "outcome_var vs predictor_var by predictor_var2")

```

### Interaction Effects

Interaction terms are predictor variables that significantly affect the relationship between the outcome variable and other predictor variables. Thus, this results in an MLR with has groups with different intercepts AND slopes. Therefore, to account for this interaction, we must multiply the predictor variables together as such:

$$
Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3(X_1 \times X_2)
$$

##### Hypothesis Testing

$$
\beta_3 = 0 \text { : No interaction effect - the variables are only additive}
$$

$$
\beta_3 \neq 0 \text { : Interaction effect - the variables either have a synergistic or interference effect}
$$

##### Example

```{r}
#| label: interact-effects
#| message: false
#| warning: false
#| echo: true  

#data$predictor_var1 <-relevel(data$predictor_var1, ref = "cat_level_1")

#interaction_model <- lm(outcome_var~predictor_var1+predictor_var2+predictor_var1*predictor_var2,data = data)

#summary(interaction_model)

```

The reference level can be changed in the following example:

```{r}
#| label: ref-level-change
#| message: false
#| warning: false
#| echo: true  

##tomhs <- tomhs %>% 
  ##mutate(wtchgcat1 = fct_relevel(wtchgcat, ##"gained"))

```

After conducting a linear regression, your reference level is the category that is the one that does not appear in your output. It is also the intercept value $\beta_0$.

#### Tests for Multi-Category Predictors

A Type III ANOVA F-Test or a Likelihood of Ratio Test can either be conducted to determine if a multi-category predictor is statistically significant.

```{r}
#| label: ANOVA-F-TEST
#| message: false
#| warning: false
#| echo: true  

#library(car)

## model <- lm(
## formula = cholestorl_change ~ weight_change_Cat + other_covariates,
## data = df)


## summary(Anova(model, type = "III"))

```

In the case of a likelihood ratio test, we are comparing the full model to the reduced model.

```{r}
#| label: likelihood-ratio-test
#| message: false
#| warning: false
#| echo: true 

#model_full <- lm(outcome_var ~ #predictor_var + other_covariates, data = data)

#model_reduced <- lm(outcome_Var ~ other_covariates, data = data)

#summary(anova(model_reduced, model_full))

```

#### Outliers & Influential Points

Outliers are measurements outside of the expected normal range. Influential points are points that affect the model slope. These two can be detected in a linear model using the following methods:

1.  Standardized Residuals \| abs(rstandard(model))\> OR \< +/-2 ⇒ observation is an outlier
2.  Students Residuals \| abs(rstudent(model)) \< 3 ⇒ outlier
3.  Cooks Distance \| cooks.distance(model) \> 1 ⇒ outlier
    1.  See @sec-MLR-diagnostics for Cooks Distance example
    2.  We can also obtain a list of Cook's Distance using the following code: cooks.distance(model_adjusted)
4.  DFFITS \| dffits(model)
    1.  If $\text{|DFFITS|} > (2 \times \sqrt (p/n)$ ⇒ influential observation
        1.  p  = number of parameters
5.  DFBETA \| dfbeta(model)
6.  PRESS \| PRESS(model)
    1.  This tells us which variables in the model may or may not be legit
    2.  Small PRESS value ⇒ Good predictability and can generalize new data
    3.  Big PRESS value ⇒ Bad predictability and may be sensitive to data in general

The code-chunk below displays a plot of DFFITS values vs observation numbers to determine influential observations.

```{r}
#| label: fig-DFFITS
#| fig-cap: "caption"
#| fig-cap-location: top
#| message: false
#| warning: false
#| echo: true

#library(olsrr)

#model_adjusted <- lm(data = data, outcome_var ~ predictor_var1 + predictor_var2)

#ols_plot_dffits(model_adjusted)

```

#### MLR Model Assumptions & Diagnostic Plots {#sec-MLR-diagnostics}

The linear regression model must meet the four LINE assumptions. Recall, that residuals refer to how the distance between the model's predictions to the actual values. Residuals are relevant because they tell us the accuracy of a model.

Please review the definitions below and how to interpret the graphs.

1.  Linearity
    1.  This refers to the direct relationship between the x and y variables. This means that for every increase in x-variable, we expect a direct increase in the y-outcome.
    2.  Review the Residuals vs Fitted plot. If the points appear "cloud-like" then the **model fit & linearity assumption** is met. We want an equal amount of points above and below the line with no distinct pattern in the points.
2.  Independence
    1.  This refers to to random sampling in the study. We want to ensure that one observation is completely random and does not influence another observation.
    2.  You actually cannot use any of the diagnostic plots to verify this assumption. You need to check the methodology to ensure that random sampling occurred for the study.
3.  Normality
    1.  This refers to how close the residuals follow the normal distribution.
    2.  Review the Q-Q residual plot. If the points are close to the diagonal line then it meets the **Normality of Errors assumption**. If it curves and strays from the line then it does not meet the assumption.
4.  Equal Variance (Homoscedasticity)
    1.  This refers to the residuals being at a **constant variance** at every level of the independent variable. In other words, we want to make sure the variance is the consistent.
    2.  Review the Residuals vs Levels plot. If the plot appears "cone-like" or a fan shape then the assumption is not met. We want to ensure that the distribution looks like a uniform "shot-gun" blast shape.

Furthermore, **outliers and influential points** can be diagnosed by reviewing the residuals vs leverage plot. Outliers are points with high y-axis values usually +/-3 and influential points are those in the upper-right or lower-right corners outside of Cook's Distance.

```{r}
#| label: fig-MLR-assumptions
#| fig-cap: "MLR Assumptions"
#| fig-cap-location: top
#| warning: false
#| message: false
#| echo: true


#par(mfrow = c(2,2))
#plot(adj_MLR)

```

## Logistic Regression

Logistic regression modeling is similar to linear regression modeling except we are interested in a binary outcome \[result can only be 0 or 1\]. Some differences from linear regression to logistic regression are the following:

-   Everything is on the log-odds scale

-   We are no longer calculating probability. We are calculating the odds-ratio of an event either being 0 or 1

    -   Therefore, the y-axis is on a scale from 0 to 1

-   Determining statistical significance means these two conditions must be met:

    -   The p-value is lower than alpha

    -   The exponent confidence interval **DOES NOT** include 1.0.

#### Libraries

Listed below in the code chunk are the libraries needed for logistic regression modeling.

```{r}
#| label: log-regression-library
#| message: false
#| warning: false
#| echo: true  

library(ResourceSelection) # for hosmer-lemeshow test
library(car) #for multicollinearity and linearity
library(performance) # Diagnostic plots for logistic regression
library(jtools)
library(lmtest)

```

#### Interpretation

Your output will be on the log-scale. Make sure to utilize a likelihood ratio test, variance inflation factor test, Hosmer-Lemeshow test, and a graph plotting the linearity of the logit values and x-variable to evaluate the logistic regression model diagnostics.

```{r}
#| label: dekaf-data-clean
#| warning: false
#| message: false
#| echo: true
#| include: false



dekaf <- dekaf %>% 
  mutate(cardiac_death = if_else(dcd_yn == 1, "Yes", "No", missing = NULL),
         DGF = if_else(dgf == 1, "Yes", "No", missing = NULL))
```

#### Example

**Questions:**

1.  How does categorical variable affect the odds of outcome variable, after adjusting for predictor variable 1?

```{r}
#| label: log-reg
#| echo: true

#rq1 <- glm(data = data, as.factor(outcome_var) ~ as.factor(categorical_var_b) + var1,         family = binomial(link = "logit"))

#summary(rq1)
```

```{r}
#| label: rq2
#| message: false
#| warning: false
#| echo: true  


#rq2 <- glm(data = data, as.factor(outcome_var) ~ var1 + as.factor(categorical_var), family = binomial(link = "logit"))

#summary(rq2)
```

**Answer:** When accounting for categorical variable, donor age, participants who had var1 have XYZ times the odds of outcome variable compared to participants who did not have var1. This is significantly associated with the odds of outcome_var (OR $\neq$ 1.00, p-value \< 0.05, CI: does not include 1.0).

##### Diagnostic Testing

We conducted Likelihood Ratio test to compare the goodness-of-fit between two models, the Variance Inflation Factor (VIF) to detect multicollinearity (e.g correlation between two predictor variables), and a Hosmer-Lemeshow test ensure a goodness-of-fit for our logistic regression model.

Respectively, the first test indicated that both aforementioned models are identical in predictive power (p-value = 1), the second test indicated that there were no multicollinearity issues since all predictor values were at 1.00 (VIF \< 5), and the third test indicated that both models were a good fit since we failed to reject the null hypothesis (p-value \> 0.05). The logit vs donor age scatterplot, was used to verify the linearity assumption. The trajectory of the blue curve is extremely close to the red line and does not diverge into any other shape, and thus, we verify that the linearity is met.

##### Conclusion

Based on the adjusted models, var1 is positive and is significantly associated with change in odds of outcome var. And, predictor variable is positive and is significantly associated with change in odds of outcome variable.

```{r}
#| label: lr-test
#| message: false
#| warning: false
#| echo: true  

#lrtest(rq1,rq2)

```

```{r}
#| label: vif
#| message: false
#| warning: false
#| echo: true

#vif(rq1)

```

```{r}
#| label: hosmer-lemeshow
#| message: false
#| warning: false
#| echo: true  


#hoslem.test(data$outcome_var, fitted(rq1))

```

# Survival Models

Survival modeling is a method to analyze "time-to-event" data and is used to calculate the probability that a specific event will occur at a certain time.

```{r}
#| label: load-f6-data
#| message: false
#| warning: false
#| echo: true
#| include: false 

f6 <- read.csv("first6.csv")

```

```{r}
#| label: f6-data-clean-1
#| message: false
#| warning: false
#| echo: true
#| include: false


f6 <- f6 %>% 
    mutate(gender = as.factor(if_else(GENDER == 1, "Male", "Female", missing = NULL)),
         race = as.factor(case_when(RACE == 1 ~ "Latino",
                                    RACE == 2 ~ "African American",
                                    RACE == 3 ~ "White",
                                    RACE == 4 ~ "Other")),
         two_class = as.factor(if_else(twoclass == 1, "Two Class", "Three Class", missing = NULL)))

```

The following libraries are required for survival modeling

```{r}
#| label: survival-model-libraries
#| message: false
#| warning: false
#| echo: true  

library(survival) #survival analysis
library(ggfortify) #survival analysis

```

### Kaplan-Meier Plots

Kaplan-Meier Plots are used to display the survival percentage of the study demographic throughout the duration of the study.

**Question**: Create a Kaplan-Meier plot of the primary endpoint, event-free survival (variables time_var and event_var), comparing the var_strat_1 group to the the var_strat_2 group.

```{r}
#| label: fig-kaplan-meier
#| fig-cap: "Kaplan Meier Plot"
#| fig-cap-location: top
#| message: false
#| warning: false
#| echo: true

#kpm <- survfit(Surv(time_var, event_var) ~ var_strat_1, data = data)

#autoplot(kpm, conf.int = FALSE, censor.shape = "X") +
  #labs(x  = "Time", y = "Survival Percentage", title = "Survival Analysis: var_1 vs var_2", color = "Exposure")


```

### Median Survival Time

Median survival time refers to the specific point in time where 50% of the patient group has died and 50% are still alive in the Kaplan-Meier Survival Curve. In most cases, you'd just need to look where at the x-axis wherever the line is at 50%. But sometimes, we cannot obtain the median survival time because the survival curve is above 50% throughout the entire follow-up period. Therefore, we cannot calculate the estimate since median survival time is not yet reached.

### Survival at Specific Times

**Question**: Estimate the event-free survival at 4 years \[e.g 48 months\] in each treatment strategy group. That is, estimate the proportion of study participants without outcome_var.

```{r}
#| label: survival-specific-event
#| message: false
#| warning: false
#| echo: true  

#summary(kpm, times =c(48)) ##note: the times argument is going to vary based on your time unit

```

**Answer**: The var_1 has a better survival estimate at four years because the survival estimate for var_1 is 76.3% whereas the survival estimate for var_2 is 71.2%.

### Log-Rank Test

A log-rank test is used to determine if there is a statistically significant difference in survival times between groups (e.g Treatment A vs Treatment B). This test gives equal weight to all time points.

**Question**: Carry out a test to determine whether event-free survival in the two treatment strategy groups is statistically different.

Is there a difference in event-free survival between the var strategy groups? Cite evidence to support your answer.

```{r}
#| label: log-rank-test
#| message: false
#| warning: false
#| echo: true  

#survdiff(Surv(time_var, event_var) ~ var_1, data =data, rho = 0) 
## make sure to set rho = 0 for log-rank

```

**Answer**: There is no difference in event-free survival between the var strategy groups (p = 0.2; p-value \> 0.05).

### Gehan-Breslow-Wilcoxon Test

This test is used to compare the survival curves of 2+ independent groups AND gives weight to events that occur early in the study.

```{r}
#| label: wilcoxon
#| message: false
#| warning: false
#| echo: true  

#survdiff(Surv(time_var, event_var) ~ var_1, data = data, rho = 1) 
## make sure to set rho = 1 for GBW

```

## Cox Proportional Hazards Model

The Cox Proportional Hazards Model is used in survival analysis to compare the relationship between survival time and a predictor variable. The output we get is the Hazard Ratio.

The predictor variable(s) can either be continuous and/or categorical.

### Hypothesis Testing for Hazard Ratios (HR)

$$
H_0: HR = 1 \text{ | There is no risk difference between groups}
$$

$$
H_1: HR < 1 \text{ | The treatment group has a lower risk of the event (better survival or fewer negative events). For example, HR = 0.5 means the event is half as likely to happen in the treatment group compared to the control group.}
$$

$$
H_1: HR > 1 \text{ | The treatment group has a higher risk of the event (worse outcome). For example, HR = 2.0 means the event is twice as likely to happen in the treatment group.}
$$

#### Data Cleaning

```{r}
#| label: clean-f6-data
#| message: false
#| warning: false
#| echo: true
#| include: false

f6 <- f6 %>% 
  mutate(Gender = as.factor(if_else(GENDER == 1, "Male", "Female", missing = NULL)),
         Race = as.factor(case_when(RACE == 1 ~ "Latino",
                          RACE == 2 ~ "African American",
                          RACE == 3 ~ "White",
                          RACE == 4 ~ "Other")),
         Above200 = as.factor(if_else(above200 == 1, "Above 200 cell count", "Below 200 cell count")),
         Randgrp = as.factor(case_when(RANDGRP == 1 ~ "PI + NRTI",
                                       RANDGRP == 2 ~ "NNRTI + NRTI",
                                       RANDGRP == 3 ~ "PI + NNRTI + NRTI")),
         twoclass = as.factor(twoclass),
         TwoClass = as.factor(
           fct_relevel(case_when(twoclass == 1 ~ "Two Class",
                                        twoclass == 2 ~ "Three Class"),"Two Class")), ## adjust Two Class as reference level
         Unit = relevel(as.factor(Above200), ref = "Below 200 cell count")) # make UNIT into factor

```

### Example

In this example, we are calculating the hazard proportion estimate of the var_f factors for outcome_var while adjusting for var_1 and var_2 variables.

To run the hazard proportion model we first need to call the following libraries:

#### Library

```{r}
#| label: lib-cox
#| message: false
#| warning: false
#| echo: true  

library(ggfortify) ## to calculate "Schoenfeld residuals vs time"
library(survminer)
library(multcomp)
library(emmeans) ## to calculate hazard ratios
```

This model compares the class_A to the Class_B strategies.

**Question:** How does the class_A strategy affect the hazard of the primary endpoint relative to the two-class strategy, after adjusting for var_1 and var_2?

```{r}
#| label: cox-model
#| message: false
#| warning: false
#| echo: true

#cox_model <- coxph(Surv(time_var,event_var) ~ var_class + var_1 + var_2, data = data)

#summary(cox_model) <-- this will be in log form. Make sure to convert to exp form.

```

#### Interpretation \[Completely made-up for reporting purposes\]

The class_A strategy has a 14.5% higher risk of hazard of event_variable than the Class_B strategy, holding all else constant. This is based on the hazard ratio of 1.145, which indicates a higher risk of hazard for the class_A strategy as it is greater than the null value. Since the confidence interval includes one (0.931, 1.409) and the p-value is greater than alpha (p-value = 0.2), there is no statistically significant difference in the Class A vs. Class B strategy after adjusting.

### Pairwise Comparisons

If we have a multi-categorical predictor, we then need to conduct pairwise comparisons using a Tukey HSD test. In this scenario, since we are only comparing two levels against each other, a Tukey HSD test is not needed.

```{r}
#| label: cox-tukey
#| message: false
#| warning: false
#| echo: true  

#cox.multcomp <- glht(cox_model, linfct = mcp(m.f = "Tukey"))

#exp(confint(cox.multcomp)$confint)

```

### Checking Proportional Hazards Assumption Example

To verify assumptions, we first need to create an adjusted model, and then we need to conduct a Schoenfeld test.

```{r}
#| label: HR-model-unadj
#| message: false
#| warning: false
#| echo: true  

#class_model_unadj <- coxph(Surv(time_var, event_var) ~ var_class, data = data)

#summary(class_model_unadj)

```

If the p-value is greater than alpha, and the $\beta$ throughout the time series is inside the boundaries, then the assumptions are met. Meaning that the model is most likely valid.

```{r}
#| label: fig-schoenfeld
#| fig-cap: "Schoenfeld Residuals vs Time Plot"
#| fig-cap-location: top
#| message: false
#| warning: false
#| echo: true

#class_resid <- cox.zph(class_model_unadj)

#ggcoxzph(class_resid)

```

## Poisson Regression Model

Poisson regression is used to model the **probability** of a number of events per a fixed amount of time. For example, it could be counting the number of cars passing a road in one hour. In other words, we are calculating the average rate of events.

```{r}
#| label: lhs-import
#| message: false
#| warning: false
#| echo: true
#| include: false

lhs <- read.csv("lhs.csv")

lhs <- lhs %>% 
  mutate(cigs5 = if_else(AV5CIGS == 0.5, 1.0, AV5CIGS),
         Race = case_when(
           race == 1 ~ "White",
           race == 2 ~ "Black",
           race == 3 ~ "Asian",
           race == 4 ~ "Native American",
           race == 5 ~ "Other"
         ),
         Gender = if_else(
           AGENDER == "M", "Males", "Females"
         ),
        Marital = case_when(
          marital0 == 1 ~ "Never Married",
          marital0 == 2 ~ "Married",
          marital0 == 3 ~ "Widowed",
          marital0 == 4 ~ "Separated",
          marital0 == 5 ~ "Divorced"
        ),
        TreatmentGroup = fct_relevel(alphagroup, "UC")
  )

```

### Example

In the example below, we are fitting a Poisson regression model to estimate the effect of the treatment group on outcome.

```{r}
#| label: poisson-reg-model
#| message: false
#| warning: false
#| echo: true  

#trt_group_model <- glm(outcome ~ treatment_var, data = data, family = poisson(link = "log"))

#summary(trt_group_model)

```

##### Interpretation

**Question:**  What is the estimated outcome rate ratio for participants in the treatment group compared to the control group?

**Answer \[made-up for interpretation purposes\]:** The estimated rate ratio for participants in the treatment group is 0.67 in comparison to the control group (95% CI: 0.66 to 0.68). Holding other factors constant, this means that the treatment group had 33% behavior / activity in comparison to the control group. This finding is statistically significant given that our p-value is lower than alpha and the 95% confidence interval does not include 1.0 (p-value \< 0.001, 95% CI: 0.66 to 0.68).

##### Model Evaluation for Overdispersion Test

To evaluate the poisson regression, we need to review the ratio of the residual deviance to its degree of freedo**ms**. **If the model ratio is super super greater than 1 then that is a sign of overdispersion.**

```{r}
#| label: overdispersion-test
#| message: false
#| warning: false
#| echo: true  

#model_ratio <- (trt_group_model$deviance / trt_group_model$df.residual)

#model_ratio

#model_ratio > 10 ## checking if model_ratio is super huge

```

### Quasi-poisson Regression Model for Overdispersed Data

Overdispersion can be dealt with either using a Quasi-Poisson Regression or a Negative Binomial Model

```{r}
#| label: quasi-poisson
#| message: false
#| warning: false
#| echo: true  

#model1.qp <- glm(outcome_var ~ treatment_var, data = data, family = quasipoisson(link = "log"))

# summary(model1.qp) <-- this will be in log form. Make sure to convert to exponent form

```

#### Interpretation

-   Question: What is the effect of predictor variable on outcome variable incidence rate? That is, describe the magnitude and direction of the effect AND its statistical significance in the context of this problem. Use evidence from the previous questions to support your answer.

-   Answer \[made-up for interpretation purposes\]: In terms of treatment variable, Group A has a 64% higher outcome incidence rate than Group B while holding all else constant (IRR = 1.64). This is statistically significant given that p-value is less than alpha and the 95% confidence interval does not contain 1.0 (p-value \< 0.001, CI: 1.48 to 1.82).

Make sure to use the exponent form when interpreting the values.

### Offset Term

```{r}
#| label: fh-load-clean
#| message: false
#| warning: false
#| echo: true
#| include: false

fh <- read.csv("Framingham_dataset.csv")

fh <- fh %>% 
  mutate(followupYR = followup / 365.25,
         CHD = as.factor(if_else(chdfate == 1, "Developed CHD", "No CHD")),
         Sex = as.factor(if_else(sex == 1, "Male", "Female")),
         Sex_women_ref = fct_relevel(Sex, "Female"))

```

```{r}
#| label: tbl-offset-term
#| tbl-cap: "Incidence Rate for Framingham Dataset"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true

library(gtExtras)

#offset_term_table <- data %>% 
 # group_by(class_var) %>% 
  #summarize(
   # outcome_cases = sum(outcome == 1),
    #outcome_py = sum(followupYR),
    #outcome_IR = ((outcome_cases / outcome_py)*100) ## CH Incidence rate calculation)  %>% 
  #gt() %>% ## convert into gt() table
  #tab_header(title = "Outcome Incidence Metrics") %>%
  #cols_label(outcome_cases = "Total Cases",
    #outcome_py = "Person-Years",
    #outcome_IR = "Incidence Rate per 100 person-years") 
  

#offset_term_table
```

An offset term is used in Poisson regression to account for observations that had more time to occur. For instance, if we do not have an offset term, our poisson regression will assume that every person will have an equal opportunity to get a disease. Whereas, in reality, a singular person may have cancer for 10 years and another person may have cancer for 2 years. In other words, the offset variable will always be the time variable (e.g follow-up), and is needed to account for the different variations in follow-up time (e.g 10 years follow-up vs 1-year; the Poisson model helps distinguish who’s more at risk) Therefore, the offset term is used to account for this time variation. The offset term is only needed if we are SPECIFICALLY looking at the RATE.

# Other Regression Models

Regression models such as Log-binomial and Ordinal do not follow the Central Limit Theorem and/or normal probability distribution.

## Log-binomial Regression Model

Log-binomial regression is used to determine the relative risk for binary outcomes. We need the following libraries for log binomial regression. The output is an odds-ratio and interpreting said output is similar to previous OR information in this document.

```{r}
#| label: log-binomial-reg
#| message: false
#| warning: false
#| echo: true  

library(tidymodels) ## generate log models via tidyverse
library(emmeans) ## calculate ratios (e.g risk ratio)
```

```{r}
#| label: dekaf-data-clean-v2
#| message: false
#| warning: false
#| echo: true
#| include: false

dekaf <- dekaf %>% 
  mutate(Gender = as.factor(if_else(gender == 1, "Male", "Female")),
         DGF = as.factor(if_else(dgf == 1, "Yes", "No")),
         Gender_women_ref = relevel(Gender, ref = "Female"))

```

### Example

**Question:** What is the effect of var1 on outcome, after adjusting for variable 2?

```{r}
#| label: tbl-logistic-reg-model
#| tbl-cap: "Logistic Regression Model"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


#logistic_reg_model <- glm(outcome_var ~ var1 + var2, data = data)

#logistic_reg_model

```

#### Interpretation \[fake-data\]

$$\text{logit}[P(\text{outcome\_var}_1 \le j)] = (\beta_1 \times \text{var1}) + (\beta_2 \times \text{var2})$$

Holding all else constant, Group A has 12% higher odds outcome risk compared to Group B (OR = 1.88). This is statistically significant given that the p-value is lower than alpha and the confidence interval does not include 1.0 (CI: 1.30 to 2.76).

## Ordinal Regression Model

Ordinal Regression is used to target cumulative ordered categories (e.g low, medium, high) where the ranking matters.

We need the following libraries for ordinal regression:

```{r}
#| label: ordinal-regression-library
#| message: false
#| warning: false
#| echo: true  

library(jtools)
library(VGAM)
```

#### Example

Question: Write the appropriate adjusted ordinal regression equation you would use to determine whether outcome variable depends on whether a person did variable 1 after adjusting for variable 2.

$$
\text{logit}[P(\text{outcome\_variable} \le j] = (\beta_1 \times \text{var1}) + (\beta_2 \times \text{var2})
$$

```{r}
#| label: tbl-adj-ordinal-regression
#| tbl-cap: "Adjusted Ordinal Regression"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true
#ordinal_regression <- vglm(outcome_var ~ var1 + var2, #data = data, family = cumulative(link = "logitlink", #parallel = TRUE))

#summary(ordinal_regression)

```

#### Interpretation \[fake data\]

-   Using the model results, what is the estimated common odds ratio for the outcome variable? 

    -   The estimated common odds-ratio for the outcome variable is 0.90. This means patients who smoke have a 10% decrease in odds of reporting a better status (e.g “Excellent”).

-   Using the model results, is there a significant relationship between doing variable 1 and outcome variable, after adjusting for variable 2? Explain your reasoning.

    -   No, after adjusting for age, we conclude that there is not a statistically significant relationship between having ever smoked cigarettes and general health at 12 months considering that the p-value is greater than alpha and the confidence interval includes 1.0 (OR = 0.90, p-value = 0.4, CI: 0.70 to 1.16).

# Sample Size Estimates and Power

Sample size and power are linked through four primary variables. If you know three, you can calculate the fourth:

1.  **Alpha (**$\alpha$): The significance level (usually 0.05). This is the risk of a **Type I error** (finding an effect that isn't actually there).

2.  **Beta (**$\beta$): The risk of a **Type II error** (failing to find an effect that actually exists).

3.  **Power (**$1 - \beta$): The probability of correctly rejecting the null hypothesis when it is false.

    1.  This is usually set to a level of 0.80 0r 0.90 unless specified otherwise.

4.  **Effect Size:** The "signal" you are looking for—how much of a difference exists between the two groups (e.g., a 10 mmHg drop in blood pressure). Represented as delta or "d".

    1.  To detect a small effect, we need a large sample

## Continuous Outcomes

#### Estimating Sample Sizes in Group Means

```{r}
#| label: tbl-sample-size-continuous
#| tbl-cap: "Estimating Sample Size from a Continuous Outcome"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true

library(pwr)

#twogrpmean.n <- pwr.t.test(d = 0.20, sig.level = 0.05, power = 0.90, type = "two.sample")

#summary(twogrpmean.n)
```

### Estimating Multiple Sample Sizes via Power Table

```{r}
#| label: tbl-sample-size-power-table
#| tbl-cap: "Estimating Multiple Sample Sizes via Power Table"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true

#candidate.n <- seq(from = 25, to = 1000, by = 25) #creates a list 25, 50, 75, 100, ...

#twogrpmean.power <- pwr.t.test(d = 0.2, sig.level = 0.05,n = candidate.n, type = "two.sample")

#summary(twogrpmean.power)
```

## **Using a Power Curve to Compare Two Different MCIDS (Effect Sizes)**

```{r}
#| label: fig-power-curve
#| fig-cap: "Using a Power Curve to Compare Two Different MCIDS (Effect Sizes)"
#| fig-cap-location: top
#| message: false
#| warning: false
#| echo: true


# creating list of sample sizes
n.values <-seq(from = 5, to = 100, by = 5)

# getting power for each sample size in list
twogrpmean.power.medium<- pwr.t.test(d = 0.5, sig.level = 0.05, n = n.values, type = "two.sample")

twogrpmean.power.large<- pwr.t.test(d = 0.8, sig.level = 0.05, n = n.values, type = "two.sample")
# plotting the first curve

plot(n.values, twogrpmean.power.medium$power, type = "b",
xlab = "sample size", ylab = "power", ylim=c(0,1))

lines(n.values, twogrpmean.power.medium$power, type = "l")

# adding the points with dashed lines of the second curve to plot
points(n.values, twogrpmean.power.large$power, type = "b", col=2, lty=2, pch = 2)
# adding a legend for each curve
legend("bottomright", c( "d = 0.5", "d = 0.8"), lty = c(1,2),pch = c(1,2), col = c(1,2), cex=1.5)


```

What if my outcome is Continuous (Group Means)?

-   Information needed to calculate sample size:

    -   Mean

    -   Standard Deviation

    -   Effect Size (also known as delta or MCID)

    -   Power Level (usually 0.80 unless stated otherwise)

## Binary Outcomes

What if my outcome is Binary (Proportions)?

-   Information needed to calculate sample size:

    -   Standard Deviations **IS NOT NEEDED**

    -   Baseline Proportion

    -   Comparison Proportion

    -   Effect Size (MCID)

#### Estimating Sample Sizes

```{r}
#| label: tbl-sample-size-grp-prop
#| tbl-cap: "Estimating Sample Size in a Group Proportion"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true

twogrpp.n <- pwr.2p.test(h = ES.h(p1 = 0.85, p2 = 0.70), sig.level = 0.05, power = 0.90)

twogrpp.n

```

## Calculating Multiple Sample Sizes with the same MCID {#sec-binary-outcome-multiple-sample-size-same-MCID}

```{r}
## Define Baseline Proportions
p1_levels <- seq(from = 0.50, to = 0.75, by = 0.05)

# Level 1: 50% to 60%
res1 <- power.prop.test(p1=0.50, p2=0.60, sig.level=0.05, power=0.80, n=NULL)
n1   <- ceiling(res1$n)

# Level 2: 55% to 65%
res2 <- power.prop.test(p1=0.55, p2=0.65, sig.level=0.05, power=0.80, n=NULL)
n2   <- ceiling(res2$n)

# Level 3: 60% to 70%
res3 <- power.prop.test(p1=0.60, p2=0.70, sig.level=0.05, power=0.80, n=NULL)
n3   <- ceiling(res3$n)

# Level 4: 65% to 75%
res4 <- power.prop.test(p1=0.65, p2=0.75, sig.level=0.05, power=0.80, n=NULL)
n4   <- ceiling(res4$n)

# Level 5: 70% to 80% (Original target)
res5 <- power.prop.test(p1=0.70, p2=0.80, sig.level=0.05, power=0.80, n=NULL)
n5   <- ceiling(res5$n)

# Level 6: 75% to 85%
res6 <- power.prop.test(p1=0.75, p2=0.85, sig.level=0.05, power=0.80, n=NULL)
n6   <- ceiling(res6$n)

## 3. Combine results into the final table
twogrpp.pwrtable <- data.frame(
  Baseline_Suppression = p1_levels,
  Comparison_Suppression = p1_levels + 0.10,
  Required_N_per_Group = c(n1, n2, n3, n4, n5, n6)
)

# Display the final table
twogrpp.pwrtable
```

## Using a Power Curve to Estimate Power based on Sample Size {#sec-binary-outcome-power-curve}

```{r}
## Define range of samples

n_range <- seq(from = 10, to = 300, by = 10)


## Get Sample Size for each proportion

pow_60 <- 
  power.prop.test(
    p1 = 0.60,
    p2 = 0.70,
    n = n_range,
  )$power

pow_70 <-
  power.prop.test(
    p1 = 0.70,
    p2 = 0.80,
    n = n_range,
  )$power

## Plot
plot(n_range, pow_70, 
     type = "b", 
     col = "blue", 
     pch = 16,
     xlab = "Sample Size (per group)", 
     ylab = "Power", 
     ylim = c(0, 1))

# Add the 60% line
lines(n_range, pow_60, type = "b", col = "red", pch = 16)

# Add a reference line for 80% power
abline(h = 0.80, lty = 2, col = "darkgrey")

## 5. Add a legend that matches your data
legend("bottomright", 
       legend = c("Baseline 70% (Target 80%)", "Baseline 60% (Target 70%)"), 
       col = c("blue", "red"), 
       pch = 16, 
       lty = 1, 
       cex = 0.8) # Adjusted cex so it fits better
```
