---
title: "Project 2: Coding Reference Sheet"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    fontsize: 0.9em
    code-fold: FALSE
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 5
    smooth-scroll: TRUE
    embed-resources: TRUE
    fig-align: center
editor: visual
---

# Exploratory Data Analysis Concepts

## Importing Data

### How do I import data using the tidyverse?

#### Library {#sec-library}

To first import data into R, we need to load the tidyverse package. This package gives us the tools needed to both import and clean data.

In the code chunk below, we load in the tidyverse package (and other packages that are more relevant in other sections).

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true

library(tidyverse) ## data cleaning 
library(kableExtra) ## make table from output via kable()
library(gtsummary) ## make table via tbl_regression()

  
```

#### Import Data

We use the read.csv() function to import .csv files. Make sure to specify the file pathway and assign it to a variable using "\<-".

In the code chunk below, we import the tomhs dataset by specifying it's location and file extension.

```{r}
#| label: import-data
#| warning: false
#| message: false
#| echo: true

tomhs <- read.csv("tomhs.csv")  


```

## Data Wrangling

### Creating New Variables {#sec-new-var}

To create new variables, we first want to overwrite our dataset with changes with the "\<-" operator.

Use the pipe operator to combine functions together in one code chunk.

Use mutate() to define our new variables.

Note: Use the if_else() function when you have a singular class, and use case_when() when the categorical variable has multiple classes.

In the code chunk below, we create two new variables, "Sex" and "Race". Sex is created by an if_else() statement and Race is created by a case_when() statement. Both variables are converted into factors - this is important for statistical analysis purposes.

```{r}
#| label: new-var
#| warning: false
#| message: false
#| echo: true

tomhs <- tomhs %>% 
  mutate(
    Sex = as.factor(if_else(sex == 1, "Male", "Female")),
    Race = as.factor(case_when(
      race == 1 ~ "White",
      race == 2 ~ "Black",
      race == 3 ~ "Asian",
      race == 4 ~ "Other")
    ),
    weight_change = wt12-wtbl,
    cholesterol_change = chol12-cholbl)
    

  
```

### Deleting & Sub-setting Observations

Deleting observations is done through the filter() function. Use this function to specify the delete observations.

| Operator |   Definition    |
|:--------:|:---------------:|
|    &     |  AND operator   |
|    \|    |   OR operator   |
|    !     |  NOT operator   |
|   \<-    | Assign variable |

: Operators

| BaseR Functions |                Definition                 |
|:---------------:|:-----------------------------------------:|
|     is.na()     | Filters to rows where there are NA values |
|                 |                                           |

: BaseR Functions

We should create a new dataset just in case we want to refer to the original dataset.

In the code chunk below, we remove observations in the age column that are equivalent to NA. And, we also subset these changes into a new dataset, "tomhs_filter" to ensure the original dataset maintains the original observation amount.

```{r}
#| label: delete-obs-sub-set
#| warning: false
#| message: false
#| echo: true

tomhs_filter <- tomhs %>% 
  filter(!is.na(age))
  

```

# Regression Modeling

## Multiple Linear Regression (MLR) {#sec-hyp-test}

MLR is when we try to predict a **continuous outcome** with either a continuous predictor and/or categorical predictor.

The null hypothesis is that all regression coefficients are equal to zero. The alternative hypothesis is that atleast one regression coefficient is not equivalent to zero.

$$
H_0: \beta_1 = \beta_2 
$$

$$
H_1: \beta_1 \neq \beta_2
$$

Unadjusted means that we do not account or adjust for other variables that may affect the outcome variable.

Adjusted means we do account for other variables that affect the outcome variable.

### Unadjusted MLR {#sec-regression-model-tips}

Example Question: What is the effect of weight change on total cholesterol change without adjusting for anything?

In the code chunk below, we conduct an MLR by using the lm() function. This function is in BaseR so do not worry about pulling an packages.

The general argument for most regression models is lm(y-variable \~ x-variable + x2-variable.... data = dataset). In other words, you're basically saying "the y-variable can be explained by the following x-variables from this dataset".

We use the tbl_regression() function to convert our results from raw data into a readable table. And, we specify the intercept = TRUE argument so we can have the intercept coefficient value.

```{r}
#| label: tbl-unadj-MLR
#| tbl-cap: "Unadjusted Multiple Linear Regression"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true

unadj_MLR <- lm(cholesterol_change ~ weight_change, data = tomhs)

unadj_MLR

unadj_MLR %>% 
  tbl_regression(
    intercept = TRUE
  )

```

#### Multiple Linear Regression Fitted Equation {#sec-MLR-equation}

$$
\text{outcome-variable} = intercept + (\beta_1 X_1) + (\beta_2 X_2) + ... 
$$

#### Interpretation {#sec-MLR-interpretation}

The intercept means when the x-variable is at zero, we expect the outcome variable to be at whatever the intercept is.

If the x-variable is a **continuous predictor**, for every 1 unit increase in the x-variable, we expect the y-variable to increase by the x-variable coefficient B1.

If the x-variable is a **categorical predictor** (e.g binary), and the x-variable is equivalent to 1, then we expect the y-variable to increase by the x-variable coefficient B1. This is a bit tricky to understand, so please see @sec-adjusted-MLR for an example.

If the p-value is less than 0.05, it is statistically significant.

### Adjusted MLR {#sec-adjusted-MLR}

```{r}
#| label: tbl-Adj-MLR
#| tbl-cap: "Adjusted Multiple Linear Regression"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true

adj_MLR <- lm(
  formula = cholesterol_change ~ weight_change + Sex,
  data = tomhs)

adj_MLR

adj_MLR %>% 
  tbl_regression(
    intercept = TRUE
  )

```

#### Interpretation {#sec-adj-MLR-interpretation}

Recall @sec-MLR-interpretation fitted equation.

According to our output, we have the following equation:

$$ \text{Total Cholesterol Change} = -0.04 + (0.52 \times \text{weight_change}) + (-2.8 \times \text{Sex}_{\text{Male}}) $$

This fitted equation represents the relationship between Total Cholesterol Change when we both account for weight change and sex.

In terms of the intercept, if both weight change is zero and sex is not male, we expect the total cholesterol change to decrease by -0.04 units.

In terms of the weight_change variable, if the weight change has increased by 1 unit, then we expect the total cholesterol change to increase by 0.52 units.

Refer to our data cleaning @sec-new-var where we coded the Sex variable where Males are equivalent to 1 and Female is equivalent to 0. Therefore, in terms of the Sex variable, if the patient is a male then we expect the total cholesterol change to decrease by -2.8 units. If the patient is female then we completely ignore the Sex variable since it is equivalent to zero.

### Interaction Effects

Interaction terms are predictor variables that significantly affect the relationship between the outcome variable and other predictor variables. Therefore, to account for this interaction, we must multiply the predictor variables together as such:

$$
Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3(X_1 \times X_2)
$$

#### Hypothesis Testing

$$
\beta_3 = 0 \text { : No interaction effect - the variables are only additive}
$$

$$
\beta_3 \neq 0 \text { : Interaction effect - the variables either have a synergistic or interference effect}
$$

#### Example

Fit a model estimating total cholesterol change (cholchg) from weight change (wtchg), current smoking status (nowsmk), and their interaction. Set those who do not currently smoke as the reference level. Share your code and output.

```{r}
#| label: tomhs-data-clean
#| message: false
#| warning: false
#| echo: false
#| include: false

tomhs <- tomhs %>% 
  mutate(wtchg = wt12 - wtbl,
         cholchg = chol12 - cholbl,
         nowsmk1 = as.factor(if_else(nowsmk == 1, "Current Smoker", "Non-current Smoker")))

```

```{r}
#| label: smk-interaction-model
#| message: false
#| warning: false
#| echo: true  

tomhs$nowsmk1 <-relevel(tomhs$nowsmk1, ref = "Non-current Smoker")


interaction_model <- lm(
  formula = cholchg~wtchg+nowsmk1+wtchg*nowsmk1,
  data = tomhs)

interaction_model

```

##### Interpretation

```{r}

#| label: interaction-model-exp
#| tbl-cap: "Interaction Model between Total Cholesterol Change and weight change and smoking status"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true


interaction_model %>% 
  tbl_regression(
    intercept =  TRUE
  )

```

From the example output, we get the following fitted equation:

$$
\text{Predicted Total Cholesterol Change} = -1.9984 + (0.6564 \times \text{weight_change}) + (-0.7335 \times \text{current_smoker}) + (-0.1328 \times \text{current_smoker*weight_change})
$$

And we interpret the fitted equation as such:

The model intercept is -1.998, which means that the estimated average cholesterol change during this study for someone in the reference (non-smoking) group with a weight change of zero would be -1.998 units. At a weight change of 0, current smokers have on average 0.73 units lower cholesterol change than non-smokers, but this difference is not statistically significant (p = 0.860). The coefficient for weight change (0.66) indicates that among non-smokers (the reference group), a 1-unit increase in weight change is associated with an estimated 0.66-unit increase in cholesterol change.

Therefore, in combination with the weight change coefficient, for each one unit change in weight for nonsmokers the total cholesterol change is 0.6564. While, for smokers it is 0.6564-0.1328=0.524.

### MLR Model Assumptions & Diagnostic Plots {#sec-LINE-assumptions}

The linear regression model must meet the four LINE assumptions.

Recall, that residuals refer to how the distance between the model's predictions to the actual values. In other words, residuals are relevant because they tell us the accuracy of a model.

Please review the definitions below and how to interpret the graphs.

1.  Linearity
    1.  This refers to the direct relationship between the x and y variables. This means that for every increase in x-variable, we expect a direct increase in the y-outcome.
    2.  Review the Residuals vs Fitted plot. If the points appear "cloud-like" then the assumption is met. We want an equal amount of points above and below the line, and appear cloud-like.
2.  Independence
    1.  This refers to to random sampling in the study. We want to ensure that one observation is completely random and does not influence another observation.
    2.  You actually cannot use any of the diagnostic plots to verify this assumption. You need to check the methodology to ensure that random sampling occurred for the study.
3.  Normality
    1.  This refers to how close the residuals follow the normal distribution.
    2.  Review the Q-Q residual plot. If the points are close to the diagonal line then it meets the assumption. If it curves and strays from the line then it does not meet the assumption.
4.  Equal Variance (Homoscedasticity)
    1.  This refers to the residuals being at a constant variance at every level of the independent variable. In other words, we want to make sure the variance is the consistent.
    2.  Review the Residuals vs Levels plot. If the plot appears "cone-like" or a fan shape then the assumption is not met. We want to ensure that the distribution looks like a uniform "shot-gun" blast shape.

To ensure that your linear regression model is meeting the four LINE assumptions:

```{r}
#| label: fig-MLR-assumptions
#| fig-cap: "MLR Assumptions"
#| fig-cap-location: top
#| warning: false
#| message: false
#| echo: true


par(mfrow = c(2,2))
plot(adj_MLR)

```

## Logistic Regression

Logistic regression modeling is similar to linear regression modeling except we are interested in a binary outcome \[result can only be 0 or 1\].

Some differences from linear regression to logistic regression are the following:

-   Everything is on the log-odds scale

-   We are no longer calculating probability. We are calculating the odds-ratio of an event either being 0 or 1

    -   Therefore, the y-axis is on a scale from 0 to 1

-   Determining statistical significance means these two conditions must be met:

    -   The p-value is lower than alpha

    -   The exponent confidence interval **DOES NOT** include 1.0.

#### Interpretation

Your output will be on the log-scale. Remember to use the log-scale values for writing the fitted equation and use the exponential values for interpreting the output. Also make sure to convert your variables to factors when applicable.

```{r}
#| label: dekaf-data-clean
#| warning: false
#| message: false
#| echo: false


dekaf <- read.csv("dekaf_dgf.csv")

dekaf <- dekaf %>% 
  mutate(cardiac_death = if_else(dcd_yn == 1, "Yes", "No", missing = NULL),
         DGF = if_else(dgf == 1, "Yes", "No", missing = NULL))



```

```{r}
#| label: log-reg

rq1 <- glm(data = dekaf,
           formula = as.factor(dgf) ~ as.factor(cardiac_death) + agedonor16,
           family = binomial(link = "logit"))

rq1
```

```{r}
#| label: tbl-log-reg-log-form
#| tbl-cap: "Logistic Regression Output [Log Form]"
#| tbl-cap-location: top

rq1 %>% ## log form
  tbl_regression(
    intercept = TRUE,
    exponentiate = FALSE
  )

```

```{r}
#| label: tbl-log-reg-exp-form
#| tbl-cap: "Logistic Regression Output [Exponent Form]"
#| tbl-cap-location: top

rq1 %>% ## exponent form
  tbl_regression(
    intercept = TRUE,
    exponentiate = FALSE
  )
```

## Cox Proportional Hazards Model

The Cox Proportional Hazards Model is used in survival analysis to compare the relationship between survival time and a predictor variable. The output we get is the Hazard Ratio.

The predictor variable(s) can either be continuous and/or categorical.

### Hypothesis Testing for Hazard Ratios (HR)

$$
H_0: HR = 1 \text{ | There is no risk difference between groups}
$$

$$
H_0: HR < 1 \text{ |  The treatment group has a lower risk of the event (better survival or fewer negative events). For example,means the event is half as likely to happen in the treatment group compared to the control group.}
$$

$$
H_0: HR > 1 \text{ | The treatment group has a higher risk of the event (worse outcome). For example, means the event is twice as likely to happen in the treatment group.}
$$

```{r}
#| label: load-f6-data
#| message: false
#| warning: false
#| echo: false
#| include: false 

f6 <- read.csv("first6.csv")

```

```{r}
#| label: clean-f6-data
#| message: false
#| warning: false
#| echo: false
#| include: false 

f6 <- f6 %>% 
  mutate(Gender = as.factor(if_else(GENDER == 1, "Male", "Female", missing = NULL)),
         Race = as.factor(case_when(RACE == 1 ~ "Latino",
                          RACE == 2 ~ "African American",
                          RACE == 3 ~ "White",
                          RACE == 4 ~ "Other")),
         Above200 = as.factor(if_else(above200 == 1, "Above 200 cell count", "Below 200 cell count")),
         Randgrp = as.factor(case_when(RANDGRP == 1 ~ "PI + NRTI",
                                       RANDGRP == 2 ~ "NNRTI + NRTI",
                                       RANDGRP == 3 ~ "PI + NNRTI + NRTI")),
         twoclass = as.factor(twoclass),
         TwoClass = as.factor(
           fct_relevel(case_when(twoclass == 1 ~ "Two Class",
                                        twoclass == 2 ~ "Three Class"),"Two Class")), ## adjust Two Class as reference level
         Unit = relevel(as.factor(Above200), ref = "Below 200 cell count")) # make UNIT into factor

```

### Example

In this example, we are calculating the hazard proportion estimate of the Three-class vs Two-class factors for AIDS outcome while adjusting for UNIT and above200 variables.

To run the hazard proportion model we first need to call the following libraries:

#### Library

```{r}
#| label: lib-cox
#| message: false
#| warning: false
#| echo: true  

library(ggfortify) ## to calculate "Schoenfeld residuals vs time"
library(survminer)
library(multcomp)
library(emmeans) ## to calculate hazard ratios


```

#### Example

This model compares the three-class to the two-class strategies.

Question: How does the three-class strategy affect the hazard of the primary endpoint relative to the two-class strategy, after adjusting for clinical unit and whether baseline CD4 cell count was above 200?

```{r}
#| label: cox-model
#| message: false
#| warning: false
#| echo: true

cox_model <- coxph(Surv(t2primary, primary) ~ factor(twoclass) + factor(UNIT) + factor(above200),
                data = f6)


```

##### Exponent Form

```{r}
#| label: tbl-cox-model-exponent
#| tbl-cap: "Two Class vs Three Class | Exponent Form"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true


cox_model %>% 
  tbl_regression(
    exponentiate = TRUE
  ) %>% 
  bold_labels() %>% 
  italicize_labels()
```

##### Log Form

```{r}
#| label: tbl-cox-model-log-form
#| tbl-cap: "Two Class vs Three Class | Log Form"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true


cox_model %>% 
  tbl_regression(
    exponentiate = FALSE
  ) %>% 
  bold_labels() %>% 
  italicize_labels()
```

#### Interpretation

The three-class strategy has a 14.5% higher risk of hazard of AIDS, death or CD4 cell count below 200 cells/mm3 than the two-class strategy, holding all else constant. This is based on the hazard ratio of 1.145, which indicates a higher risk of hazard for the three-class strategy as it is greater than the null value. Since the confidence interval includes one (0.931, 1.409), there is no statistically significant difference in the three-class vs. two-class strategy after adjusting.

### Pairwise Comparisons

If we have a multi-categorical predictor, we then need to conduct pairwise comparisons using a Tukey HSD test.

In this scenario, since we are only comparing two levels against each other, a Tukey HSD test is not needed.

```{r}
#| label: cox-tukey
#| message: false
#| warning: false
#| echo: true  


#cox.multcomp <- glht(AIDS_model, linfct = mcp(m.f = "Tukey"))

#exp(confint(cox.multcomp)$confint)

```

$$
\text{HR} > 1 = \text{Increased Risk} 
$$

$$
\text{HR} < 1 = \text{Decreased Risk}
$$

### Checking Proportional Hazards Assumption Example

To verify assumptions, we first need to create an adjusted model, and then we need to conduct a Schoenfeld test

```{r}
#| label: aids-model-unadj
#| message: false
#| warning: false
#| echo: true  

AIDS_model_unadj <- coxph(Surv(t2qoi, qoi) ~ TwoClass, data = f6)

summary(AIDS_model_unadj)

```

If the p-value is less than alpha and the Beta throughout the time series is inside the boundaries, then the assumptions are met.

```{r}
#| label: fig-schoenfeld
#| fig-cap: "Schoenfeld Residuals vs Time Plot"
#| fig-cap-location: top
#| message: false
#| warning: false
#| echo: true

AIDS_resid <- cox.zph(AIDS_model_unadj)

ggcoxzph(AIDS_resid)

```

## Poisson Regression Model

Poisson regression is used to model the **probability** of a number of events per a fixed amount of time. For example, it could be counting the number of cars passing a road in one hour.

In other words, we are calculating the average rate of events.

```{r}
#| label: lhs-import
#| message: false
#| warning: false
#| echo: false
#| include: false

lhs <- read.csv("lhs.csv")

lhs <- lhs %>% 
  mutate(cigs5 = if_else(AV5CIGS == 0.5, 1.0, AV5CIGS),
         Race = case_when(
           race == 1 ~ "White",
           race == 2 ~ "Black",
           race == 3 ~ "Asian",
           race == 4 ~ "Native American",
           race == 5 ~ "Other"
         ),
         Gender = if_else(
           AGENDER == "M", "Males", "Females"
         ),
        Marital = case_when(
          marital0 == 1 ~ "Never Married",
          marital0 == 2 ~ "Married",
          marital0 == 3 ~ "Widowed",
          marital0 == 4 ~ "Separated",
          marital0 == 5 ~ "Divorced"
        ),
        TreatmentGroup = fct_relevel(alphagroup, "UC")
  )

```

### Example

In the example below, we are modeling the effect of treatment group to determine the effect of treatment group on cigarette smoking at 5 years.

```{r}
#| label: cigs-model
#| message: false
#| warning: false
#| echo: true  

trt_group_cigs5_model <- glm(cigs5 ~ TreatmentGroup, 
                             data = lhs,
                             family = poisson(link = "log"))

trt_group_cigs5_model 

```

#### Exponent Form

```{r}
#| label: tbl-cigs-model-exp
#| tbl-cap: "caption"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true

trt_group_cigs5_model %>% ## this is the exponentiated form
  tbl_regression(
    intercept = TRUE,
    exponentiate = TRUE,
    label = list(TreatmentGroup ~ "Treatment Group")
  )

```

#### Log Form

```{r}
#| label: tbl-cigs-model-log-form
#| tbl-cap: "caption"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true

trt_group_cigs5_model %>% ## this is the expected log form
  tbl_regression(
    intercept = TRUE,
    exponentiate = FALSE,
    label = list(TreatmentGroup ~ "Treatment Group")
  )

```

### Model Evaluation for Overdispersion Test

To evaluate the poisson regression, we need to review the ratio of the **residual deviance** to its **degree of freedoms**. **If the model ratio is super super greater than 1 then that is a sign of overdispersion.**

```{r}
#| label: overdispersion-test
#| message: false
#| warning: false
#| echo: true  
#| 
model_ratio <- (trt_group_cigs5_model$deviance
/ trt_group_cigs5_model$df.residual)

model_ratio

model_ratio > 10

```

### Quasi-poisson Regression Model for Overdispersed Data

Overdispersion can be dealt with either using a Quasi-Poisson Regression or a Negative Binomial Model

```{r}
#| label: quasi-poisson
#| message: false
#| warning: false
#| echo: true  

model1.qp <- glm(cigs5 ~ TreatmentGroup,
                 data = lhs,
                 family = quasipoisson(link = "log"))

model1.qp

```

#### Exponent Form

```{r}
#| label: quasi-poisson-exp
#| message: false
#| warning: false
#| echo: true


model1.qp %>%  ## exponentiated form
  tbl_regression(
    exponentiate = TRUE,
    label = (TreatmentGroup ~ "Treatment Group")
  )

```

#### Log Form

```{r}
#| label: quasi-poisson-log-form
#| message: false
#| warning: false
#| echo: true


model1.qp %>%  ## expected log form
    tbl_regression(
    exponentiate = FALSE,
    label = (TreatmentGroup ~ "Treatment Group")
  )

```

### Interpretation

-   Question: Provide an answer to the research question for this section: What is the effect of sex on coronary heart disease (CHD) incidence rate? That is, describe the magnitude and direction of the effect AND its statistical significance in the context of this problem. Use evidence from the previous questions to support your answer.

-   Answer: In terms of sex, males have a 64% higher CHD incidence rate than females while holding all else constant (IRR = 1.64). This is statistically significant given that p-value is less than alpha and the 95% confidence interval does not contain 1.0 (p-value \< 0.001, CI: 1.48 to 1.82).

Make sure to use the exponent form when interpreting the values.

# Other Regression Models

Regression models such as Log-binomial and Ordinal do not follow the Central Limit Theorem and/or normal probability distribution.

## Log-binomial Regression Model

Log-binomial regression is used to determine the relative risk for binary outcomes.

We need the following libraries for log binomial regression.

```{r}
#| label: log-binomial-reg
#| message: false
#| warning: false
#| echo: true  

library(tidymodels) ## generate log models via tidyverse
library(emmeans) ## calculate ratios (e.g risk ratio)
```

```{r}
#| label: dekaf-data-clean-v2
#| message: false
#| warning: false
#| echo: false
#| include: false

dekaf <- dekaf %>% 
  mutate(Gender = as.factor(if_else(gender == 1, "Male", "Female")),
         DGF = as.factor(if_else(dgf == 1, "Yes", "No")),
         Gender_women_ref = relevel(Gender, ref = "Female"))

```

### Example

**Questions**

-   What is the effect of having ever smoked cigarettes on general health at 12 months, after adjusting for age?

-   What is the effect of age on general health at 12 months, after adjusting for having ever smoked cigarettes?

```{r}
#| label: tbl-logistic-reg-model
#| tbl-cap: "Logistic Regression Model: DGF Outcome by Gender"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


model_1 <- logistic_reg() %>% 
  set_engine("glm") %>% 
  set_mode("classification") %>% 
  fit(DGF ~ Gender, data = dekaf_dgf)

tidy(model_1) ## this is to make the output readable



```

```{r}

#| label: tbl-logistic-reg-model-exp
#| tbl-cap: "Logistic Regression Model: DGF Outcome by Gender"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true

model_1 %>% ## exponent form
  tbl_regression(
    intercept = TRUE,
    exponent = TRUE
  ) %>% 
  bold_labels() %>% 
  italicize_labels()


```

```{r}

#| label: tbl-logistic-reg-model-log-form
#| tbl-cap: "Logistic Regression Model: DGF Outcome by Gender"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true

model_1 %>% ## log form
  tbl_regression(
    intercept = TRUE,
    exponent = FALSE
  ) %>% 
  bold_labels() %>% 
  italicize_labels()

```

#### Interpretation 

$$log(DGF) = -1.6 + (0.63\times \text{Gender}_{\text{male}}) $$

Holding all else constant, men have a 12% higher odds DGF risk compared to women (OR = 1.88). This is statistically significant given that the p-value is lower than alpha and the confidence interval does not include 1.0 (CI: 1.30 to 2.76).

## Ordinal Regression Model

Ordinal Regression is used to target cumulative ordered categories (e.g low, medium, high) where the ranking matters.

We need the following libraries for ordinal regression:

```{r}
#| label: ordinal-regression-library
#| message: false
#| warning: false
#| echo: true  

library(jtools)
library(VGAM)

```

#### Example

Question: Write the appropriate adjusted ordinal regression equation you would use to determine whether general health at 12 months (ql12_1) depends on whether a person ever smoked after adjusting for age. Use the actual names of the variables (not just X or Y) and beta0, beta1, etc. for the coefficients.

$$
\text{logit}[P(\text{ql_12}_1 \le j] = (\beta_1 \times \text{eversmk}) + (\beta_2 \times \text{age})
$$

```{r}
#| label: tbl-adj-ordinal-regression
#| tbl-cap: "Adjusted Ordinal Regression"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true


ordinal_regression <- vglm(
                formula = GenHealth ~ Eversmk + age,
                data = tomhs,
                family = cumulative(link = "logitlink", parallel = TRUE))

ordinal_regression

```

```{r}

#| label: tbl-adj-ordinal-regression-log-form
#| tbl-cap: "Adjusted Ordinal Regression Log Form"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true

model_1 %>%  ## log form
  tbl_regression(
    intercept = TRUE,
    exponentiate = FALSE
    ) %>% 
  bold_labels() %>% 
  italicize_levels()

```

```{r}
#| label: tbl-adj-ordinal-regression-log-form
#| tbl-cap: "Adjusted Ordinal Regression Exponent Form"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true

model_1 %>%  ## exponent form
  tbl_regression(
    intercept = TRUE,
    exponentiate = TRUE,
    ) %>% 
  bold_labels() %>% 
  italicize_levels() 

```

-   Using the model results, what is the estimated common odds ratio for the eversmk variable? Provide an interpretation of this common odds ratio in the context of the problem. 

    -   The estimated common odds-ratio for the eversmk variable is 0.90. This means patients who smoke have a 10% decrease in odds of reporting a better health status (e.g “Excellent”).

-   Using the model results, is there a significant relationship between having ever smoked cigarettes and general health at 12 months, after adjusting for age? Explain your reasoning.

    -   No, after adjusting for age, we conclude that there is not a statistically significant relationship between having ever smoked cigarettes and general health at 12 months considering that the p-value is greater than alpha and the confidence interval includes 1.0 (OR = 0.90, p-value = 0.4, CI: 0.70 to 1.16).

-   Using the model results, what is the estimated common odds ratio for the age variable? Provide an interpretation of this common odds ratio in the context of the problem.

    -   The estimated common-odds ratio for the age variable is 0.98. This indicates that for every one-year increase in age, the odds of a patient reporting a better general health status (e.g., ‘Excellent’ vs. all lower categories) decrease by 2%, holding smoking status constant.

-   Using the model results, is there a significant relationship between age and general health at 12 months, after adjusting for having ever smoked cigarettes? Explain your reasoning.

    -   After adjusting for having ever smoked cigarettes, we conclude that the relationship between age and general health at 12 months is not statistically significant given that the p-value is greater than alpha and the confidence interval includes 1.0 (p-value = 0.12, CI: 0.97 to 1.00).

# Sample Size Estimates and Power

## Continuous Outcomes

## Binary Outcomes

# .qmd Notes
