---
title: "PROPORTIONAL HAZARDS REGRESSION"
author: "Patrick Torralba"
format: html
editor: visual
output: bookdown:html_document2
---

# PART 3: Carrying out and Interpreting Analyses

## library + load + dataclean

```{r}
#| label: library
#| include: false

library(tidyverse)
library(survival)
library(ggfortify)
library(survminer)
library(multcomp)
library(captioner) # for figures
```

```{r fig label}

fig_label <- captioner(prefix = "Figure")

```
* Note: use the fig_label function **that we created** to automatically label
** captioner is **NOT** a function we can j use. We must create a function using captioner()

```{r}
#| label: load datasets
#| include: false

f6 <- read.csv("first6.csv")

f6_twoclass <- read.csv("first6_twoclass.csv")

```

```{r}
#| label: dataclean

f6 <- f6 %>% 
  mutate(twoclass = as.factor(twoclass),
         primary = as.numeric(as.character(primary)))

```

## MODEL 1: Adjusted Cox proportional hazards model comparing three-class to two-class strategies

```{r}
#| label: Cox hazard model

model1 <- coxph(Surv(t2primary, primary) ~ twoclass, data = f6)

summary(model1)

```

-   Note: inside the Surv() function, the first argument is the time variable and the second argument is the event indicator
-   the output of the first line (after n = ), shows us the regression coefficient estimates in **LOG FORM** \*\* **the second line gives us the exp() values**

### Proportional Hazards Assumption Test

```{r}
#| label:  Schoenfeld residuals vs time

model1_resid <- cox.zph(model1)

model1_title <- fig_label("plot1", "Model 1 Assumption: Schoenfeld residuals vs time")

ggcoxzph(model1_resid, caption = model1_title) 


```

-   Note: To interpret the graph, look to see if there is a time trend \*\* proportional hazard assumptions met if **there is lack of a time trend** \*\* also make sure to look at p-value. high-pvalue means assumption is met

## MODEL 2: Comparing the NNRTI to PI strategies by age subgroups

```{r}
#| label: create indicator variable

f6_twoclass <- f6_twoclass %>% 
  mutate(agege40 = as.factor(if_else(AGE >= 40, 1, 0)))

```
* Note: 1 = 40+ yo, 0 = < 40 yo

```{r indicator plot}
#| fig-cap: "Age Distribution"

ggplot(f6_twoclass, aes(x = agege40, fill = agege40)) +
  geom_bar() +
  labs(x = NULL,
       y = "Count",
       title = fig_label("plot2","Age Distribution: Below/Above 40 Years Old")) 

```

