---
title: "PROPORTIONAL HAZARDS REGRESSION"
author: "Patrick Torralba"
format: html
editor: visual
---

# PART 3: Carrying out and Interpreting Analyses

## library + load + dataclean
```{r}
#| label: library

library(tidyverse)
library(survival)
library(ggfortify)
library(survminer)
library(multcomp)
```


```{r}
#| label: load dataset

f6 <- read.csv("first6.csv")

```


```{r}
#| label: dataclean

f6 <- f6 %>% 
  mutate(twoclass = as.factor(twoclass),
         primary = as.numeric(as.character(primary)))

```


## MODEL 1: Adjusted Cox proportional hazards model comparing three-class to two-class strategies

```{r}
#| label: Cox hazard model

model1 <- coxph(Surv(t2primary, primary) ~ twoclass, data = f6)

summary(model1)

```
* Note: inside the Surv() function, the first argument is the time variable and the second argument is the event indicator
* the output of the first line (after n = ), shows us the regression coefficient estimates in **LOG FORM**
** **the second line gives us the exp() values**


### Proportional Hazards Assumption Test
```{r}
#| label:  Schoenfeld residuals vs time

model1_resid <- cox.zph(model1)

ggcoxzph(model1_resid)

```

* Note: To interpret the graph, look to see if there is a time trend
** proportional hazard assumptions met if **there is lack of a time trend**
** also make sure to look at p-value. high-pvalue means assumption is met
