---
title: "PROPORTIONAL HAZARDS REGRESSION"
author: "Patrick Torralba"
format: 
  html:
    fig-cap-location: top
editor: visual
---

# PART 3: Carrying out and Interpreting Analyses

## library + load + dataclean

```{r}
#| label: library
#| results: 'hide'

library(tidyverse)
library(survival)
library(ggfortify) ## to calculate "Schoenfeld residuals vs time"
library(survminer)
library(multcomp)
library(emmeans) ## to calculate hazard ratios
```

```{r}
#| label: load datasets
#| include: false

f6 <- read.csv("first6.csv")

f6_twoclass <- read.csv("first6_twoclass.csv")

```

```{r}
#| label: dataclean

f6 <- f6 %>% 
  mutate(twoclass = as.factor(twoclass),
         primary = as.numeric(as.character(primary)))

```

## MODEL 1: Adjusted Cox proportional hazards model comparing three-class to two-class strategies

```{r}
#| label: Cox hazard model

model1 <- coxph(Surv(t2primary, primary) ~ twoclass, data = f6)

summary(model1)

```

-   Note: inside the Surv() function, the first argument is the time variable and the second argument is the event indicator
-   the output of the first line (after n = ), shows us the regression coefficient estimates in **LOG FORM**
    -   **the second line gives us the exp() values**

### Proportional Hazards Assumption Test

```{r}
#| label:  fig-schoenfeld-residuals
#| fig-cap: "Schoenfeld residuals vs time"
#| fig-path: Exports/

model1_resid <- cox.zph(model1)

ggcoxzph(model1_resid) 


```

-   Note: To interpret the graph, look to see if there is a time trend
-   proportional hazard assumptions met if **there is lack of a time trend**
-   also make sure to look at p-value. high-pvalue means assumption is met

## MODEL 2: Comparing the NNRTI to PI strategies by age subgroups

```{r}
#| label: create indicator variable

f6_twoclass <- f6_twoclass %>% 
  mutate(agege40 = as.factor(if_else(AGE >= 40, 1, 0)))

```

-   Note: 1 = 40+ yo, 0 = \< 40 yo

```{r indicator plot}
#| label: fig-age-dist
#| fig-cap: "Age Distribution: Above/Below 40 Years Old"
#| fig-path: Exports/

ggplot(f6_twoclass, aes(x = agege40, fill = agege40)) +
  geom_bar() +
  labs(x = NULL,
       y = "Count")

```

-   Note: to add figure captions, I should use both label and fig-cap arguments at the top
    -   To make it look more professional, I can remove the title and j add it to the fig-cap argument
    -   The fig-path argument j specifies to R where the image should be downloaded to

```{r 17}

model_2 <- coxph(Surv(t2primary, primary) ~ twoclass * agege40 + UNIT + above200, data = f6_twoclass)

summary(model_2)
```

-   Note: first argument = time variable
    -   2nd argument = outcome event
    -   since this question asks if treatment differs by age. I should calculate the interaction effect.
    -   to adjust for additional variables, j add them to the equation
-   The above equation is for multiple Cox Regression variables. Make sure to convert the x variable(s) into a factorial.

## Estimate hazard ratio

```{r 21}

hr_age <- emmeans(model_2, pairwise ~ twoclass | agege40, type = "response")

hr_age
```

-   Note: To estimate the hazard ratio, I need to consider the interaction effect.
    -   the above equation is how to estimate the hazard ratios
-   How to read output:
    -   Output is already in normal form and **NOT IN LOG FORM**
        -   Meaning that they are ready to be interpreted from the get-go.
    -   response = Hazard Ratio estimate
    -   asymp.LCL / asymp.UCL = Lower / Upper Confidence Level bounds
-   Remember for this specific example, two-class is set to only those in the two-class group.
