---
title: "Week 8: Homework - Proportional Hazards Regression"
author: "Patrick Torralba"
format: 
  html:
    number-sections: true
    fig-cap-location: top
editor: visual
---

Proportional Hazards Regression Report:

Research Question: How does the three-class strategy affect the hazard of AIDS relative to the two-class strategy, after adjusting for clinical unit and whether baseline CD4 cell count was above 200?

## Library

```{r}
#| code-fold: true
#| code-summary: "Click here to view loaded libraries"
#| message: false
#| warning: false

library(tidyverse)
library(survival)
library(ggfortify) ## to calculate "Schoenfeld residuals vs time"
library(survminer)
library(multcomp)
library(emmeans) ## to calculate hazard ratios
library(gtsummary) ## to make table
library(flextable) ## export tables to Word
```

## load data

```{r}

f6 <- read.csv("first6.csv")
```

## data cleaning

```{r}
#| code-fold: true
#| code-summary: "Click here to view data cleaning code"

f6 <- f6 %>% 
  mutate(Gender = as.factor(if_else(GENDER == 1, "Male", "Female", missing = NULL)),
         Race = as.factor(case_when(RACE == 1 ~ "Latino",
                          RACE == 2 ~ "African American",
                          RACE == 3 ~ "White",
                          RACE == 4 ~ "Other")),
         Above200 = as.factor(if_else(above200 == 1, "Above 200 cell count", "Below 200 cell count")),
         Randgrp = as.factor(case_when(RANDGRP == 1 ~ "PI + NRTI",
                                       RANDGRP == 2 ~ "NNRTI + NRTI",
                                       RANDGRP == 3 ~ "PI + NNRTI + NRTI")),
         twoclass = as.factor(twoclass),
         TwoClass = as.factor(
           fct_relevel(case_when(twoclass == 1 ~ "Two Class",
                                        twoclass == 2 ~ "Three Class"),"Two Class"))) ## adjust Two Class as reference level

```

-   Note: added code fold option so that I can view this better in html

# Exploratory Data Analysis

### Table 1: Study Population

```{r}
#| label: tbl-study-pop
#| tbl-cap: "Study Population Characteristics"
#| tbl-cap-location: top

t1 <- f6 %>% 
  dplyr::select(AGE, Gender, Race, Above200, Randgrp) %>% 
  tbl_summary(by = Randgrp,
              label = list("AGE" = "Age",
                             "Above200" = "Cell Count")) %>% 
  add_overall() %>% 
  as_flex_table()

t1
```

-   Note to self: I must type "tbl-..." exactly so R knows that the code chunk is explicitly referring to a table

```{r export table 1 to word}
#| code-fold: true
#| code-summary: "Click here to view table 1 export"
#| eval: false


t1 %>% 
  save_as_docx(path = "Exports/Table_1.docx")

```

-   Note to self: Variables I should pull

    -   AGE

    -   GENDER

    -   RACE

    -   above200

    -   RANDGRP

### Kaplan-Meier plot

```{r km model}

AIDS_fit <- survfit(Surv(t2qoi, qoi) ~ TwoClass, data = f6)

```

```{r}
#| label: fig-kmplot
#| fig-cap: "Kaplan-Meier: AIDS Survival Analysis"
#| warning: false

autoplot(AIDS_fit) +
  xlab("Time (Months)") +
  ylab("AIDS Survival Percentage") +
  labs(title = "Time to AIDS",
       subtitle = "By Treatment Type") +
  theme_minimal()

```

# Results

### Multiple Cox Proportional Hazards Regression

Note to self: I am calculating the hazard proportion estimate of the Three-class vs Two-class factors regarding AIDS while adjusting for UNIT + above200

```{r}

AIDS_model <- coxph(Surv(t2qoi, qoi) ~ TwoClass + UNIT + Above200, data = f6)

summary(AIDS_model)
```

### Pairwise comparisons for multi-categorical predictors

```{r}
#| eval: false

### I don't think this is necessary because we are only comparing two levels against each other.

cox.multcomp <- glht(AIDS_model, linfct = mcp(m.f = "Tukey"))

exp(confint(cox.multcomp)$confint)
```

### Create AIDS Table

```{r}
#| label: tbl-AIDS
#| tbl-cap: "Multiple Cox Proportional Hazards Regression"
#| tbl-cap-location: top

AIDS_table <- tbl_regression(AIDS_model, exponentiate = TRUE) %>%
  bold_labels() %>%
  italicize_levels() %>% 
  as_flex_table()  ## this is so I can copy-paste directly from the html render
  

AIDS_table ## turns out I can copy-paste directly from the html render!!! 
```

### Checking proportional hazards assumptions: Schoenfeld residuals vs time plot

```{r}
#| label: fig-schoenfeld
#| fig-cap: " Schoenfeld Residuals vs Time Plot"
#| fig-cap-location: top


AIDS_resid <- cox.zph(AIDS_fit)

ggcoxzph(AIDS_resid)

```

The residuals were created by plotting the @fig-kmplot model values.
