---
title: "Log Binomial Regression"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    self_contained: true
    embed-resources: true
    toc: true
    smooth-scroll: true
    number-sections: true
editor: visual
---

# Library + Load Data

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true

library(tidyverse)
library(epitools) ## Make 2x2 tables
library(gtsummary) ## convert results into table
library(broom) ## convert baseR results into tibbles
library(tidymodels) ## generate log models via tidyverse

```

```{r}
#| label: load data
#| warning: false
#| message: false
#| echo: false

dekaf_dgf <- read.csv("dekaf_dgf (1).csv")

```

# Data Clean

```{r}
#| label: dekaf-dgf-df-clean

dekaf_dgf <- dekaf_dgf %>% 
  mutate(Gender = as.factor(if_else(gender == 1, "Male", "Female")),
         DGF = as.factor(if_else(dgf == 1, "Yes", "No")),
         Gender_women_ref = fct_relevel(Gender, "Female"))



```

# 2x2 Table: Gender by DGF

Note: tbl_cross() is used to make really good 2x2 tables in html via the gtsummary package

```{r}
#| label: tbl-gender-dgf
#| tbl-cap: "2x2 Table: Gender by DGF"
#| tbl-cap-location: top


table_gender_dgf <- dekaf_dgf %>% 
  tbl_cross( 
    row = Gender,
    col = DGF) %>% 
  bold_labels()
    
table_gender_dgf
  

```

Note: tbl_uvregression() can calculate the odds-ratio in an html table.

```{r}
#| label: tbl-gender-dgf-odds-ratio
#| tbl-cap: "2x2 Table: Gender by DGF Odds-Ratio"
#| tbl-cap-location: top
#| eval: !expr packageVersion("dplyr") >= "1.2.0"
#| warning: false
#| message: false
#| echo: true


dekaf_dgf %>% 
  dplyr::select(Gender, DGF) %>% 
  tbl_uvregression(
    method = glm,
    y = DGF,
    method.args = list(family = binomial),
    exponentiate = TRUE
  ) %>% 
  bold_labels() %>% 
  italicize_levels() 



```

# Model 1: Logistic model for DGF by recipient gender {#sec-logistic-reg-ex}

```{r}
#| label: tbl-logistic-reg-model
#| tbl-cap: "Logistic Regression Model: DGF Outcome by Gender"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


model_1 <- logistic_reg() %>% 
  set_engine("glm") %>% 
  set_mode("classification") %>% 
  fit(DGF ~ Gender, data = dekaf_dgf)

tidy(model_1) ## this is to make the output readable


model_1 %>% ## exponent form
  tbl_regression(
    intercept = TRUE,
    exponent = TRUE
  ) %>% 
  bold_labels() %>% 
  italicize_labels()

model_1 %>% ## log form
  tbl_regression(
    intercept = TRUE,
    exponent = FALSE
  ) %>% 
  bold_labels() %>% 
  italicize_labels()

```

## Model 1: Fitted Equation {#sec-fitted-equation-ex}

Notes: \* When writing the fitted equation, make sure to use the log form. \* To copy-paste this from an html file into google docs, copy the code as a TeX command and insert this into google docs as an equation \* Or, download the LaTeX editor Chrome Extension and j use this in gdocs. But this will be an image instead.

$$log(DGF) = -1.6 + (0.63\times \text{Gender}_{\text{male}}) $$

::: hidden
-   I can use the two dollar sign characters to make equations via LaTex syntax
:::

# Logistic Regression Notes

-   Question: Provide an answer to the research question for this section: How does gender affect the odds of DGF? Remember to interpret the magnitude and direction of the estimate and its significance. Use evidence from the previous questions to support your answer.

-   Answer: Holding all else constant, men have a 12% higher odds DGF risk compared to women (OR = 1.88). This is statistically significant given that the p-value is lower than alpha and the confidence interval does not include 1.0 (CI: 1.30 to 2.76).

-   @sec-logistic-reg-ex

-   @sec-fitted-equation-ex

# Model 2: Log-binomial model for DGF by recipient gender

```{r}
#| label: tbl-log-binomial-model
#| tbl-cap: "caption"
#| tbl-cap-location: top

```

