---
title: "Log Binomial Regression"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    self_contained: true
    embed-resources: true
    toc: true
    smooth-scroll: true
    number-sections: true
editor: visual
---

# Library + Load Data

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true

library(tidyverse)
library(epitools) ## Make 2x2 tables
library(gtsummary) ## convert results into table
library(broom) ## convert baseR results into tibbles
library(tidymodels) ## generate log models via tidyverse

```

```{r}
#| label: load data
#| warning: false
#| message: false
#| echo: false

dekaf_dgf <- read.csv("dekaf_dgf (1).csv")

```

# Data Clean

```{r}
#| label: dekaf-dgf-df-clean

dekaf_dgf <- dekaf_dgf %>% 
  mutate(Gender = as.factor(if_else(gender == 1, "Male", "Female")),
         DGF = as.factor(if_else(dgf == 1, "Yes", "No")),
         Gender_women_ref = fct_relevel(Gender, "Female"))



```

# 2x2 Table: Gender by DGF

Note: tbl_cross() is used to make really good 2x2 tables in html via the gtsummary package

```{r}
#| label: tbl-gender-dgf
#| tbl-cap: "2x2 Table: Gender by DGF"
#| tbl-cap-location: top


table_gender_dgf <- dekaf_dgf %>% 
  tbl_cross( 
    row = Gender,
    col = DGF) %>% 
  bold_labels()
    
table_gender_dgf
  

```

Note: tbl_uvregression() can calculate the odds-ratio in an html table.

```{r}
#| label: tbl-gender-dgf-odds-ratio
#| tbl-cap: "2x2 Table: Gender by DGF Odds-Ratio"
#| tbl-cap-location: top
#| eval: !expr packageVersion("dplyr") >= "1.2.0"
#| warning: false
#| message: false
#| echo: true


dekaf_dgf %>% 
  dplyr::select(Gender, DGF) %>% 
  tbl_uvregression(
    method = glm,
    y = DGF,
    method.args = list(family = binomial),
    exponentiate = TRUE
  ) %>% 
  bold_labels() %>% 
  italicize_levels() 



```

# Model 1: Logistic model for DGF by recipient gender {#sec-logistic-reg-ex}

```{r}
#| label: tbl-logistic-reg-model
#| tbl-cap: "Logistic Regression Model: DGF Outcome by Gender"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


model_1 <- logistic_reg() %>% 
  set_engine("glm") %>% 
  set_mode("classification") %>% 
  fit(DGF ~ Gender, data = dekaf_dgf)

tidy(model_1) ## this is to make the output readable


model_1 %>% ## exponent form
  tbl_regression(
    intercept = TRUE,
    exponent = TRUE
  ) %>% 
  bold_labels() %>% 
  italicize_labels()

model_1 %>% ## log form
  tbl_regression(
    intercept = TRUE,
    exponent = FALSE
  ) %>% 
  bold_labels() %>% 
  italicize_labels()

```

## Model 1: Fitted Equation {#sec-fitted-equation-ex}

When writing the fitted equation, make sure to use the log form.

$$log(DGF) = -1.6 + (0.63\times \text{Gender}_{\text{male}}) $$

::: hidden
-   I can use the "\$\$" characters to make equations via LaTex syntax
:::
