---
title: "Log Binomial Regression"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    self_contained: true
    embed-resources: true
    toc: true
    smooth-scroll: true
    number-sections: true
editor: visual
---

# Library + Load Data

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true

library(tidyverse)
library(epitools) ## Make 2x2 tables
library(gtsummary) ## convert results into table
library(broom) ## convert baseR results into tibbles
library(tidymodels) ## generate log models via tidyverse
library(emmeans) ## calculate ratios (e.g risk ratio)
library(gt) ## convert results to table

```

```{r}
#| label: load data
#| warning: false
#| message: false
#| echo: false

dekaf_dgf <- read.csv("dekaf_dgf (1).csv")

```

# Data Clean

```{r}
#| label: dekaf-dgf-df-clean

dekaf_dgf <- dekaf_dgf %>% 
  mutate(Gender = as.factor(if_else(gender == 1, "Male", "Female")),
         DGF = as.factor(if_else(dgf == 1, "Yes", "No")),
         Gender_women_ref = relevel(Gender, ref = "Female"))



```

# 2x2 Table: Gender by DGF

Note: tbl_cross() is used to make really good 2x2 tables in html via the gtsummary package

```{r}
#| label: tbl-gender-dgf
#| tbl-cap: "2x2 Table: Gender by DGF"
#| tbl-cap-location: top


table_gender_dgf <- dekaf_dgf %>% 
  tbl_cross( 
    row = Gender,
    col = DGF) %>% 
  bold_labels()
    
table_gender_dgf
  

```

Note: tbl_uvregression() can calculate the odds-ratio in an html table.

```{r}
#| label: tbl-gender-dgf-odds-ratio
#| tbl-cap: "2x2 Table: Gender by DGF Odds-Ratio"
#| tbl-cap-location: top
#| eval: !expr packageVersion("dplyr") >= "1.2.0"
#| warning: false
#| message: false
#| echo: true


dekaf_dgf %>% 
  dplyr::select(Gender, DGF) %>% 
  tbl_uvregression(
    method = glm,
    y = DGF,
    method.args = list(family = binomial),
    exponentiate = TRUE
  ) %>% 
  bold_labels() %>% 
  italicize_levels() 



```

# Model 1: Logistic model for DGF by recipient gender {#sec-logistic-reg-ex}

```{r}
#| label: tbl-logistic-reg-model
#| tbl-cap: "Logistic Regression Model: DGF Outcome by Gender"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


model_1 <- logistic_reg() %>% 
  set_engine("glm") %>% 
  set_mode("classification") %>% 
  fit(DGF ~ Gender, data = dekaf_dgf)

tidy(model_1) ## this is to make the output readable


model_1 %>% ## exponent form
  tbl_regression(
    intercept = TRUE,
    exponent = TRUE
  ) %>% 
  bold_labels() %>% 
  italicize_labels()

model_1 %>% ## log form
  tbl_regression(
    intercept = TRUE,
    exponent = FALSE
  ) %>% 
  bold_labels() %>% 
  italicize_labels()

```

## Model 1: Fitted Equation {#sec-fitted-equation-ex}

Notes:

-   When writing the fitted equation, make sure to use the log form.

-   To copy-paste this from an html file into google docs, copy the code as a TeX command and insert this into google docs as an equation

    -   Or, download the LaTeX editor Chrome Extension and j use this in gdocs. But this will be an image instead.

$$log(DGF) = -1.6 + (0.63\times \text{Gender}_{\text{male}}) $$

::: hidden
-   I can use the two dollar sign characters to make equations via LaTex syntax
:::

# Logistic Regression Notes

-   Question: Provide an answer to the research question for this section: How does gender affect the odds of DGF? Remember to interpret the magnitude and direction of the estimate and its significance. Use evidence from the previous questions to support your answer.

-   Answer: Holding all else constant, men have a 12% higher odds DGF risk compared to women (OR = 1.88). This is statistically significant given that the p-value is lower than alpha and the confidence interval does not include 1.0 (CI: 1.30 to 2.76).

-   @sec-logistic-reg-ex

-   @sec-fitted-equation-ex

# Model 2: Log-binomial model for DGF by recipient gender {#sec-log-binomial-model}

How does gender affect the risk of DGF?

```{r}
#| label: tbl-log-binomial-model-dgf-gender
#| tbl-cap: "Log-binomial model for DGF by recipient gender"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


model_2 <- linear_reg() %>%  ## use linear_reg most of the time
  set_engine("glm", family = binomial(link = "log")) %>% 
  fit(dgf ~ Gender_women_ref, data = dekaf_dgf)

model_2 %>% 
  tidy()

model_2 %>% 
  tbl_regression(
    label = list(Gender_women_ref = "Gender"),
    intercept = TRUE,
    exponentiate = TRUE) %>% 
  modify_header(
    estimate = "**(Exp) RR Estimate**"
  ) %>% 
  bold_labels() %>% 
  italicize_levels()

model_2 %>% 
  tbl_regression(
    label = list(Gender_women_ref = "Gender"),
    intercept = TRUE,
    exponentiate = FALSE) %>% 
  modify_header(
    estimate = "**(Log) RR Estimate**"
  ) %>% 
  bold_labels() %>% 
  italicize_levels()

```

## Log-Binomial Fitted Equation

* Remember to use the log values when writing the fitted equation!
* We are **NOT** calculating the binary outcome. 
  * We are calculating the probability the outcome occurs. Therefore, we should include log(P(DGF))

$$log(P(DGF)) = -1.8 + (0.49\times \text{Gender}_{\text{male}}) $$

## Calculate Predicted Probability from the Fitted Equation

Using the fitted equation from above, calculate the predicted probability (risk) of DGF in men, the risk of DGF in women, and the relative risk for DGF comparing men to women.  

**Note**: Whenever a question asks us to do actual calculations, we must undo the log via doing the math THEN exponentiating it
  * You cannot j look at the exponentiated table.
    * I mean we can but in the log-scale we add the numbers, whereas in the exponentiated scale, we multiply the numbers.
      * TLDR: j do the math in log-form first then exponentiate bc it's alot easier to calculate and interpret


```{r}
#| label: model-2-predicted-probability-by-hand

DGF_risk_men <- exp(-1.8+0.49) ## look at the fitted equation and undo the log

DGF_risk_women <- exp(-1.8)

DGF_RR_men_women <- round(DGF_risk_men / DGF_risk_women,2)

DGF_RR_men_women


## this is how to double check calculations by hand
```


```{r}
#| label: tbl-model-2-risks-calculations-automated
#| tbl-cap: "Model 2 Relative Risk Calculations"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true



model_2_risks <- emmeans(model_2$fit, specs = ~ Gender_women_ref, type = "response")

model_2_risks %>% 
  as.data.frame() %>% 
  dplyr::select(-SE,-df) %>% 
  gt() %>% 
  tab_header(
    title = "Model 2: Predicted DGF Risk",
    subtitle = "Analysis of Risks by Gender"
  ) %>% 
  cols_label(
    Gender_women_ref = "Gender",
    prob = "Risk Estimate",
    asymp.LCL	 = "Lower 95% CI",
    asymp.UCL = "Upper 95% CI"
  ) %>% 
  fmt_number(
    columns = is.numeric,
    decimals = 2
  )


```



```{r}

#| label: tbl-model-2-risk-ratio-calculation-automated
#| tbl-cap: "Model 2 Relative Risk Ratio Calculation"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true



model_2_risk_ratio <- contrast(model_2_risks, method = "revpairwise")

model_2_risk_ratio

model_2_risk_ratio %>% 
  as.data.frame() %>% 
  select(contrast, ratio, p.value) %>% 
  gt() %>% 
  tab_header(
    title = "DGF Relative Risk Ratio comparing Men to Women"
  ) %>% 
  cols_label(
    contrast = "Gender",
    ratio = "Risk Ratio",
    p.value = "p-value"
  ) %>% 
  fmt_number( ## format risk ratio to 2 decimals
    columns = ratio,
    decimals = 2
  ) %>% 
  fmt_number( #3 format p-value to 4 decimal places
    columns = p.value,
    decimals = 4
  )

```

The risk probability of DGF in men is 
