---
title: "Week 6: Homework - Logistic Regression"
author: "Patrick Torralba"
format: html
editor: visual
---

# Report

## Load and import data

```{r load library, include=FALSE, echo=FALSE}
library(tidyverse)
library(broom) #for aesthetic clean tables
library(flextable) # export tables
library(ResourceSelection) # for hosmer-lemeshow test
library(car) #for multicollinearity and linearity
library(gtsummary)
library(performance) # Diagnostic plots for logistic regression
library(see)
library(DHARMa) #Makes it easier to see QQ plot
```

```{r import data, include=FALSE, echo=FALSE}
url <- "https://raw.githubusercontent.com/torra034-svg/biostats/refs/heads/main/Biostats%202/Data%20Stuff/dekaf_dgf.csv"

dekaf <- read.csv(url)
```

```{r remove url, echo=FALSE}

rm(url)

```

## Dataset Data Cleaning

```{r create Cardiac Death variable}

dekaf <- dekaf %>% 
  mutate(cardiac_death = if_else(dcd_yn == 1, "Yes", "No", missing = NULL),
         DGF = if_else(dgf == 1, "Yes", "No", missing = NULL))

```

## Exploratory Data Analysis

### Table 1: Participant Demographics

```{r table 1 - data cleaning}
table_1 <- dekaf %>% 
  mutate(Race = if_else(raceblack == 1, "African American", "Other", missing = NULL),
         Gender = if_else(gender == 1, "Male", "Female", missing = NULL),
         Delayed_Graft_Function = if_else(dgf == 1, "Yes", "No", missing = NULL)) %>% 
  dplyr::select(Race, Gender, Delayed_Graft_Function, cardiac_death, bmi_base, agedonor16)

```

```{r table 1 - table summary}
table_1 <- table_1 %>% 
  tbl_summary(by = Delayed_Graft_Function,
              label = list(cardiac_death ~ "Donation After Cardiac Death",
                           bmi_base ~ "Recipient BMI at baseline (kg/m2)",
                           agedonor16 ~ "Donor age (years beyond 16, i.e. 17=1)"
                           )) %>% 
  modify_caption(caption = "**Table 1. Participant Demographics**") %>% 
  modify_header(label ~ "**Delayed Graft Function**") %>% 
   add_overall()

  
table_1
```

```{r table 1 export, eval=FALSE}

table_1 %>%
  as_flex_table() %>%
  save_as_docx(path = "Exports/Table_1_Week6.docx")
```

### Figure 1: Relationship between Donor Age and DGF

```{r graph summary of donor age vs DGF}
ggplot(dekaf, aes(x = DGF, y = agedonor16, fill = DGF)) +
  geom_boxplot() +
  labs(fill = "Delayed Graft Function",
       x = "",
       y = "Donor Age",
       title = "Figure 1. Distribution of Donor Age and Delayed Graft Function",
       subtitle = "Donor age refers to years beyond 16 (i.e. 17=1)") 


```

### Figure 2: Relationship between Cardiac Death and DGF

```{r graph summary of Cardiac Death vs DGF}

ggplot(dekaf, aes(fill = DGF, x = cardiac_death)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Donation after Cardiac Death",
       y = "Percentage",
       fill = "Delayed Graft Function",
       title = "Figure 2. Proportion of Delayed Graft Function by Cardiac Death Donation Status")

```

## Results

### Table 2

```{r logistic regression model - cardiac death + donor age}

rq1 <- glm(data = dekaf,
           as.factor(dgf) ~ as.factor(cardiac_death) + agedonor16,
           family = binomial(link = "logit"))

```

```{r rq1 odds-ratio calculation}

summ(rq1, confint = TRUE, ci.width = 0.95, digits = 4, exp = TRUE)

```

```{r tidy_rq1 export, eval=FALSE}

tidy_rq1 %>% 
  write.csv(file = "C:/Users/Lenovo/OneDrive/Documents/githubb/biostats/Biostats 2/Exports/HW4/Table2.csv")

```

### Table 3

```{r logistic regression model - cardiac death + donor age}
rq2 <- glm(data = dekaf,
           as.factor(dgf) ~ agedonor16 + as.factor(cardiac_death),
           family = binomial(link = "logit"))

```


```{r rq2 summ}

summ(rq2, confint = TRUE, ci.width = 0.95, digits = 4, exp = TRUE)

```


```{r tidy_rq2 export, eval=FALSE}

tidy_rq2 %>% 
  write.csv(file = "C:/Users/Lenovo/OneDrive/Documents/githubb/biostats/Biostats 2/Exports/HW4/Table3.csv", row.names = FALSE)

```

### Diagnostic Checks

-   Note: for a logistic regression model, you cannot use the linear diagnostic plot functions.
-   You must use the **Box-Tidwell** test to verify against the logit values.
-   Highly recommend using the check_model() function because it provides a binned residual plot for logistic regression model

####RQ1

```{r rq1 diagnostic plots}


```

```{r rq1 hoslem.test}

hoslem.test(dekaf$dgf, fitted(rq1))

```

####RQ2

```{r rq2 diagnostic plots}


```

```{r rq2 hoslem.test}

hoslem.test(dekaf$dgf, fitted(rq2))

```
