---
title: "Week 11: Homework"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    code-fold: FALSE
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 3
    smooth-scroll: TRUE
    embed-resources: TRUE
    number-sections: TRUE
editor: visual
---

# Set-up

## Library

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true

library(tidyverse)
library(gtsummary) ## make table 1
library(psych) ## descriptive stats for table 2
library(broom) ## clean object output 
library(kableExtra) ## convert object outputs into table
library(stringr)
library(tidymodels)
library(jtools) ## needed for ordinal regression calculation
library(VGAM) ## needed for ordinal regression calculation  

```

## Load Data

```{r}
#| label: load data
#| message: false
#| warning: false
#| echo: false

tomhs <- read.csv("tomhs (2).csv")

```

## Data Clean

```{r}
#| label: data-clean
#| warning: false
#| message: false
#| echo: true


tomhs <- tomhs %>% 
  mutate(Sex = if_else(sex == 1, "Male", "Female"),
         Race = case_when(race == 1 ~ "White",
                          race == 2 ~ "Black",
                          race == 3 ~ "Asian",
                          race == 4 ~ "Other"),
         Income = case_when(income == 1 ~ "Less than $9,999",
                            income == 2 ~ "$10,000 to $19,999",
                            income == 3 ~ "$20,000 to $29,999",
                            income == 4 ~ "$30,000 to $39,999",
                            income == 5 ~ "$40,000 to $49,999",
                            income == 6 ~ "$50,000 to $59,999",
                            income == 7 ~ "$60,000 to $69,999",
                            income == 8 ~ "$70,000 or more"),
         Marital = case_when(marital == 1 ~ "Never married",
                             marital == 2 ~ "Separated",
                             marital == 3 ~ "Divorced",
                             marital == 4 ~ "Widowed",
                             marital == 5 ~ "Married"),
         Education = case_when(educ == 1 ~"Eighth grade or less",
                               educ == 2 ~"Trade school or business school instead of high school",
                               educ == 3 ~"Some high school",
                               educ == 4 ~"High school graduate", 
                               educ == 5 ~"Trade school or business school after graduating from high school",
                               educ == 6 ~"Some college including 2 year degrees",
                               educ == 7 ~"Received bachelor's degree",
                               educ == 8 ~"Graduate or professional education beyond the bachelor's degree",
                               educ == 9 ~ "Graduate or professional degree (Specify):"),
         Eversmk = if_else(eversmk == 1, "Yes", "No"),
         "Ql12_4" = 
           fct_relevel(
             case_when(
             ql12_4 == 1 ~ "Completely satisfied",
             ql12_4 == 2 ~ "Very satisfied",
             ql12_4 == 3 ~ "Somewhat satisfied",
             ql12_4 == 4 ~ "Somewhat dissatisfied",
             ql12_4 == 5 ~ "Very dissatisfied",
             ql12_4 == 6 ~ "Completely dissatisfied"
             )
  )
  )
         
  
```

# Exploratory Data Analysis

## Participant Demographics
```{r}
#| label: tbl-study-demographics
#| tbl-cap: "Participant Demographics by Ever Smoke Status"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


tbl_1 <- tomhs %>% 
  select(age, Sex, Race, Income, Marital, Education, Eversmk, Ql12_4) %>% 
  tbl_summary(
    by = Eversmk,
    label = list(
      age = "Age",
      Ql12_4 = "Satisfaction with physical ability at 12 months",
      Eversmk = "Ever Smoked"
      )
  ) %>% 
  add_overall() %>% 
  bold_labels() %>% 
  italicize_levels()

tbl_1


```


## Ever Smoke vs Satisfaction
```{r}
#| label: fig-eversmk-vs-ql12_4
#| fig-cap: "The relationship between ever smoke status and satisfaction with physical ability at 12 months"
#| fig-cap-location: top
#| warning: false
#| message: false
#| echo: true

tomhs %>% 
  arrange(Ql12_4) %>% 
ggplot(aes(x = Eversmk, fill = Ql12_4)) +
  geom_bar(position = "fill") +
  labs(x = "Ever Smoke Status",
       y = "Proportion",
       fill = "Satisfaction Score") +
  theme_classic()

```


## Age vs Satisfaction
```{r}
#| label: fig-age-vs-ql12_4
#| fig-cap: "The relationship between age and satisfaction with physical ability at 12 months"
#| fig-cap-location: top
#| warning: false
#| message: false
#| echo: true


ggplot(tomhs, aes(x = cut(age, breaks = seq(40, 75, by = 5)), fill = Ql12_4)) +
  geom_bar(position = "fill") +
  labs(x = "Age Group", 
       y = "Proportion",
       fill = "Satisfaction Score") +
  theme_classic()


```

# Results 

## Ordinal Regression Model: The effect of Ever Smoke on Satisfaction after Adjusting for Age

```{r}
#| label: model-1-Eversmk-vs-Ql12_4-adj-age
#| warning: false
#| message: false
#| echo: true

model_1 <- vglm(
  formula = Ql12_4 ~ Eversmk + age,
  data = tomhs,
  family = cumulative(link = "logitlink", parallel = TRUE)
)

model_1


```


### Log Form

```{r}
#| label: tbl-model-1-log
#| tbl-cap: "The Effect of Ever Smoke on Satisfaction after Adjusting for Age [log]"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true

model_1 %>% 
  tbl_regression(
    intercept = TRUE,
    exponentiate = FALSE
  ) %>% 
  bold_labels() %>% 
  italicize_levels()


```


### Exponent Form

```{r}
#| label: tbl-model-1-exponent
#| tbl-cap: "The Effect of Ever Smoke on Satisfaction after Adjusting for Age [exponent]"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true

model_1 %>% 
  tbl_regression(
    intercept = TRUE,
    exponentiate = TRUE
  ) %>% 
  bold_labels() %>% 
  italicize_levels()


```

## Ordinal Regression Model: The effect of Age on Satisfaction after Adjusting for Ever Smoke

Important Note: Looks like both models produce the same output. Therefore, I don't think I need to use another model. Need to talk to Benny about this.

```{r}
#| label: model-2-age-vs-Ql12_4-adj-EverSmk
#| warning: false
#| message: false
#| echo: true

model_1 <- vglm(
  formula = Ql12_4 ~ age + Eversmk,
  data = tomhs,
  family = cumulative(link = "logitlink", parallel = TRUE)
)

model_1 %>% 
  tbl_regression(
    intercept = TRUE
  )


```
