---
title: "Collaborative Key 2"
author: "Patrick Torralba"
date: "2025-09-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load library
```{r}
library(psych)
```



#import data
```{r}
study <- read.csv(file.choose(), header =  TRUE)

```

#examine variable data types
```{r}
str(study)
```

#examine DAScore and AverageSleep
```{r}
str(study$DASScore)

str(study$AverageSleep)
```

#Summary Statistics: Average Sleep
```{r}
summary(study$AverageSleep)
IQR(study$AverageSleep)
describe(study$AverageSleep)
```

#Summary Statistics: DAScore
```{r}
summary(study$DASScore)
IQR(study$DASScore)
describe(study$DASScore)
```

#Histogram for DAScore vairable
hist(study$DASScore,
 xlab = "DAS Score",
 main = "DAS Score Frequency")


#Boxplot for DasScore variable
```{r}
boxplot(study$DASScore,
        main = "DAS Score Distribution",
        xlab = "DAS Score",
        ylab = "Frequency")
```

#AverageSleep Boxplot
```{r}
boxplot(study$AverageSleep,
        main = "AverageSleep Boxplot",
        xlab = "Average sleep")
```

#Potential Outlier for Average Sleep?
##Lower fence: Q1 - (1.5*IQR)
```{r}
7 - (1.5*21)
```


#AverageSleep Histogram Modified
```{r}
hist(study$AverageSleep,
     breaks = 15)
```

#AverageSleep Histogram Modified | Lots of Bins
```{r}
hist(study$AverageSleep,
     breaks = 50)
```

#AverageSleep Histogram Modified | Lots of Bins
```{r}
hist(study$AverageSleep,
     breaks = 5)
```