---
title: "Project 1"
author: "Patrick Torralba"
format: html
editor: visual
---

```{r}
#| label: library
#| echo: false
#| include: false
#| warning: false

library(tidyverse) # general
library(gtsummary) # create tables
library(flextable) # convert to flex table for copy-paste purposes
library(jtools) # logistic regression
library(ResourceSelection) # logistic regression
library(gtsummary) # make models into object for best copy-pasting
library(lmtest) # Likelihood ratio test 
library(car) # for vif and multicollinearity
```

```{r}
#| label: data
#| echo: false

surgery_timing <- read.csv("Surgery Timing.csv")

```

## Data cleaning

```{r}
#| label: data cleaning

surgery_timing <- surgery_timing %>% 
  mutate(Gender = as.factor(if_else(gender == 1, "Male", "Female")),
         Race = as.factor(case_when(
                          race == 1 ~ "Caucasian",
                          race == 2 ~ "African American",
                          race == 3 ~ "Other")),
         Asa_status = as.factor(case_when(
                                asa_status == 1 ~ "ASA 1-2",
                                asa_status == 2 ~ "ASA 3",
                                asa_status == 3 ~ "ASA 4-6")),
         Complication = factor(if_else(complication == 0, "No Complication", "Complication"),
                               levels = c("No Complication", "Complication")),
         Dow = as.factor(case_when(dow == 1 ~ "Monday",
                         dow == 2 ~ "Tuesday",
                         dow == 3 ~ "Wednesday",
                         dow == 4 ~ "Thursday",
                         dow == 5 ~ "Friday" )),
         Dow = fct_relevel(Dow, "Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
  

```

```{r}
#| label: surgery_timing_1

# this dataset does not have any missing values. This is needed for the likelihood ratio test 

surgery_analysis <- surgery_timing %>% 
  dplyr::select(Complication, hour, Dow, age, Gender, bmi, Asa_status) %>% 
  na.omit()
```


## Exploratory Data Analysis

::: {.content-hidden}

### Variables I should pull for table 1:
    -   age
    -   race
    -   gender
    -   bmi
    -   asa_status
        -   this refers to a patients physical health and combordities
    -   mort30
    
:::

### Table - Participant Demographics
```{r}
#| label: tbl-participant-demo
#| tbl-cap: "Participant Demographics"
#| tbl-cap-location: top
#| warning: false

table_1 <- surgery_timing %>% 
  tbl_summary(by = Complication,
              include = Gender:Complication,
              label = "Asa_status" ~ "ASA Status") %>% 
  add_overall() %>% 
  modify_footnote(update = label ~ "ASA: American Society of Anesthesiologists physical status classification") %>% 
  as_flex_table()

table_1
```



### Figure - Surgeries and Complications by Hour Histogram
```{r}
#| label: fig-hist-hour-comp
#| fig-cap: "Surgeries and Complications by Hour"
#| fig-cap-location: top


ggplot(surgery_timing, aes(x = hour,  fill = Complication)) +
  geom_histogram(binwidth = 1, position = "dodge") +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  theme_minimal() +
  theme(legend.position = "top") + 
  labs(x = "Hour (24-Hour Time)",
       y = "Surgery Count",
       fill = "Patient Status")


```

### Figure - Complications by Day of Week 
```{r}
#| label: fig-comp-dow
#| fig-cap: "Surgical In-hospital Complications by Day of Week"
#| fig-cap-location: top


ggplot(surgery_timing, aes(x = Dow, y = complication, fill = )) +
  geom_col() +
  labs(x = NULL,
       y = "Complication Count")

## I do not include the title because I screenshot the figure-cap instead

```

## Results

We are doing a logistic regression because the outcome is binary (complication vs no complication).

 
### Hour vs Complication model

Full Model
```{r}
#| label: hour-comp-model-full

hour_comp_model_full <- glm(Complication ~ hour + age + Gender + bmi + Asa_status, data = surgery_analysis, family = binomial(link = "logit"))

```


Reduced Model
```{r}
#| label: hour-comp-model-reduced

hour_comp_model_reduced <- glm(Complication ~ hour, data = surgery_analysis, family = binomial(link = "logit"))



```

Logistic Regression Analysis

```{r}
#| label: tbl-hour-comp-model
#| tbl-cap: "Complication by Hour"
#| tbl-cap-location: "top"

tbl_regression(hour_comp_model_full, exponentiate = TRUE) %>% 
  bold_labels()

```

### Dow vs Complication model

Dow Full Model
```{r}
#| label: Dow-comp-model-full

Dow_comp_model_full <- glm(Complication ~ Dow + age + Gender + bmi + Asa_status, data = surgery_analysis, family = binomial(link = "logit"))

```


Dow Reduced Model
```{r}
#| label: dow-comp-model-reduced

Dow_comp_model_reduced <- glm(Complication ~ Dow, data = surgery_analysis, family = binomial(link = "logit"))

```


Dow Logistic Regression
```{r}
#| label: tbl-dow-comp-model
#| tbl-cap: "Complication by Day of Week"
#| tbl-cap-location: "top"


tbl_regression(Dow_comp_model_full, exponentiate = TRUE) %>% 
  bold_labels()

```

## Model Diagnostics

### Hour vs Comp Diagnostics

Likelihood Ratio Test
```{r}
#| label: Hr vs Comp- Likelihood Ratio Test

lrtest(hour_comp_model_reduced, hour_comp_model_full) ## look for different diagnostic. Look at w6 software lesson.

```

Variance Inflation Factor to check for Multicollinearity
```{r}
#| label: Hr vs Comp - Variance Inflation Factor

vif(hour_comp_model_full)

```


Linearity Assumption

```{r}
#| label: subset hour vs comp

hour_comp_data <- subset(surgery_timing, select = c(age, bmi, hour))

hour_comp_predictors <- colnames(hour_comp_data)

hour_comp_data$logit <- predict(hour_comp_model_full)



```

```{r}
#| label: fig-hr-comp-linearity
#| fig-cap: "Linearity Assumption Check for Hour vs Complication Model"
#| fig-cap-location: top

pivot_longer(hour_comp_data, names_to = "predictors", values_to = "predictor.value", -logit) %>% 
  ggplot(aes(x = predictor.value, y = logit)) +
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess", color = "blue", se = FALSE) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  facet_wrap(facets = ~ predictors, scales = "free_x") +
  theme_bw()

```

### Dow vs Comp Diagnostics

Likelihood Ratio Test
```{r}
#| label: Dow vs Comp - Likelihood Ratio Test

lrtest(Dow_comp_model_reduced, Dow_comp_model_full)
```


Variance Inflation Factor to check for Multicollinearity
```{r}
#| label: Dow vs Comp - VIF

```


Linearity Assumption
```{r}
#| label: subset - dow vs comp

```

```{r}
#| label: fig-dow-comp-linearity
#| fig-cap: "Day of Week Linearity Assumption"
#| fig-cap-location: top

```

