---
title: "Ordinal Regression"
author: "Patrick Torralba"
format:
  html:
    html-math-method: mathjax
    theme: flatly
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 3
    smooth-scroll: TRUE
    embed-resources: TRUE
    number-sections: TRUE
editor: visual
---

# Setup

## Library

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true

library(tidyverse)
library(gtsummary) ## make table 1
library(psych) ## descriptive stats for table 2
library(broom) ## clean object output 
library(kableExtra) ## convert object outputs into table
library(stringr)
library(tidymodels)
library(multilevelmod) ## tidymodels extension for proportional_odds
library(latex2exp) ## write latex equations in plots and tables
library(jtools) ## needed for ordinal regression calculation
library(VGAM) ## needed for ordinal regression calculation

```

## Load Data

```{r}
#| label: load data
#| message: false
#| warning: false
#| echo: false

tomhs <- read.csv("tomhs (2).csv")

```

## Data Clean  {#sec-data-clean}

```{r}
#| label: df-clean
#| warning: false
#| message: false
#| echo: true

tomhs <- tomhs %>% 
  mutate(
    Sex = as.factor(if_else(sex == 1, "Male", "Female")),
    Race = as.factor(case_when(
      race == 1 ~ "White",
      race == 2 ~ "Black",
      race == 3 ~ "Asian",
      race == 4 ~ "Other"
    )),
    Eversmk = as.factor(if_else(eversmk == 1, "Yes", "No")),
    GenHealth = as.factor(case_when(
      ql12_1 == 1 ~ "Excellent",
      ql12_1 == 2 ~ "Very good",
      ql12_1 == 3 ~ "Good",
      ql12_1 == 4 ~ "Fair",
      ql12_1 == 5 ~ "Poor"
    )
    )
  )

tomhs <- tomhs %>% 
  mutate(GenHealth = ordered(fct_relevel(GenHealth,
    "Excellent",
    "Very good",
    "Good",
    "Fair",
    "Poor"
  )))

```

# Exploratory Data Analysis

## Table 1: Particpant Demographics

```{r}
#| label: tbl-pt-demographic
#| tbl-cap: "Participant Demographics by Ever Smoke Status"
#| tbl-cap-location: top

tbl_1 <- tomhs %>% 
  select(age, Sex, Race, Eversmk, GenHealth) %>% 
  tbl_summary(
    by = Eversmk,
    label = list(
      Eversmk = "Ever Smoke",
      age = "Age",
      GenHealth = "General Health"
    )
  ) %>% 
  add_overall() %>% 
  bold_labels() %>% 
  italicize_levels()

tbl_1


```

## General Health Variable Table

```{r}
#| label: tbl-gen-health-var
#| tbl-cap: "General Health Variable Summary"
#| tbl-cap-location: top

tbl_2 <- tomhs %>% 
  select(ql12_1) %>% 
  describe(IQR = TRUE) %>% 
  as.data.frame() %>% 
  select(-trimmed) %>% 
  kable(digits = 2,
        col.names = tools::toTitleCase(names(.)),
        align = "c") %>% 
  kable_styling(
    bootstrap_options = c("striped","hover","condensed"),
    full_width = FALSE,
    position = "center")

tbl_2

```

## General Health Variable Distribution

```{r}
#| label: fig-gen-health-distribution
#| fig-cap: "General Health Variable Distribution"
#| fig-cap-location: top


ggplot(tomhs,aes(x = GenHealth)) +
  geom_bar() +
  labs(
    x = "General Health Status",
    y = "Count",
    title = "General Health Status Count"
  )

```

```{r}
#| label: tbl-gen-health-category-count
#| tbl-cap: "General Health Category Count"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true

tbl_3 <- tomhs %>% 
  select(GenHealth) %>% 
  group_by(GenHealth) %>%
  summarize(Count = n()) %>% 
  arrange(GenHealth) %>% 
  kable(col.names = c("General Health", "Count")) %>% 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  )
  

tbl_3


```

## Relationship between General Health and Ever Smoke

```{r}
#| label: fig-genhealth-vs-eversmk
#| fig-cap: "General Health by Ever Smoke Status"
#| fig-cap-location: top
#| warning: false
#| message: false
#| echo: true


ggplot(tomhs, aes(x = GenHealth, fill = Eversmk)) +
  geom_bar() +
  labs(
    x = "General Health Status",
    y = "Count",
    fill = "Ever Smoke Status",
    title = "General Health Status by Ever Smoke Status"
  )



```

# Model 3: Adjusted Ordinal Regression

## Fitted Equation {#sec-ord-reg-fit-eq}

Write the appropriate adjusted ordinal regression equation you would use to determine whether general health at 12 months (ql12_1) depends on whether a person ever smoked after adjusting for age. Use the actual names of the variables (not just X or Y) and beta0, beta1, etc. for the coefficients.

$$
\text{logit}[P(ql12\_1\le j)] = (\beta_1 \times \text{eversmk}) + (\beta_2 \times age) 
$$

## Ordinal Regression Model: The effect of eversmk on ql12_1 after adjusting for age {#sec-ord-reg-calc-baseR}

```{r}
#| label: ordinal-reg-model-building
#| warning: false
#| message: false
#| echo: true


model_1 <- vglm(
                formula = GenHealth ~ Eversmk + age,
                data = tomhs,
                family = cumulative(link = "logitlink", parallel = TRUE))

model_1
```

```{r}
#| label: tbl-ord-reg-output
#| tbl-cap: "Results: The Effect of Smoking on General Health after Adjusting for Age [log]"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


model_1 %>%  ## log form
  tbl_regression(
    intercept = TRUE,
    exponentiate = FALSE
    ) %>% 
  bold_labels() %>% 
  italicize_levels()


```

```{r}
#| label: tbl-ord-reg-output-exponent
#| tbl-cap: "Results: The Effect of Smoking on General Health after Adjusting for Age [exponent]"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


model_1 %>%  ## exponent form
  tbl_regression(
    intercept = TRUE,
    exponentiate = TRUE,
    ) %>% 
  bold_labels() %>% 
  italicize_levels() 

```

### Ordinal Regression Model Interpretation  {#sec-ordinal-model-reg-interpret}

-   Using the model results, what is the estimated common odds ratio for the eversmk variable? Provide an interpretation of this common odds ratio in the context of the problem. 

    -   The estimated common odds-ratio for the eversmk variable is 0.90. This means patients who smoke have a 10% decrease in odds of reporting a better health status (e.g “Excellent”).

-   Using the model results, is there a significant relationship between having ever smoked cigarettes and general health at 12 months, after adjusting for age? Explain your reasoning.

    -   No, after adjusting for age, we conclude that there is not a statistically significant relationship between having ever smoked cigarettes and general health at 12 months considering that the p-value is greater than alpha and the confidence interval includes 1.0 (OR = 0.90, p-value = 0.4, CI: 0.70 to 1.16).

-   Using the model results, what is the estimated common odds ratio for the age variable? Provide an interpretation of this common odds ratio in the context of the problem.

    -   The estimated common-odds ratio for the age variable is 0.98. This indicates that for every one-year increase in age, the odds of a patient reporting a better general health status (e.g., 'Excellent' vs. all lower categories) decrease by 2%, holding smoking status constant.

-   Using the model results, is there a significant relationship between age and general health at 12 months, after adjusting for having ever smoked cigarettes? Explain your reasoning.

    -   After adjusting for having ever smoked cigarettes, we conclude that the relationship between age and general health at 12 months is not statistically significant given that the p-value is greater than alpha and the confidence interval includes 1.0 (p-value = 0.12, CI: 0.97 to 1.00).

# Ordinal Regression Summary Notes 

1.  Data Cleaning tips
    1.  Make sure to set your categorical variable into an ordered factor
        1.  This is super important for generating the model
    2.  @sec-data-clean
2.  How do I write the fitted equation?
    1.  Make sure to write the fitted equation **ONLY IN LOG FORM**
    2.  @sec-ord-reg-fit-eq
3.  How do I generate an ordinal regression model?
    1.  @sec-ord-reg-calc-baseR
4.  How do I interpret the output?
    1.  Make sure to interpret the output **ONLY IN EXPONENT FORM**
    2.  Make sure to also state "a X% increase/decrease in the odds" instead of "more/less likely"
        1.  Remember. When we are dealing with odds-ratios we must not use probability vocab
    3.  @sec-ordinal-model-reg-interpret
5.  I don't understand. What does "logitlink(P\[Y\<=1\])" mean in the table output?
    1.  This represents the order of the categorical variable
        1.  In this scenario, 1 represents Excellent, 2 represents Very Good.
        2.  This all depends on how you ordered your factored categorical variable
    2.  @sec-ord-reg-calc-baseR
