---
title: "Lesson 3 Lab"
author: "Patrick Torralba"
date: "2025-09-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

#Load library
```{r}
library(epiDisplay)
```

#Confidence Intervals for a Mean in R

#Import Data
```{r}
lhs<-read.csv(file = "C:/Users/Lenovo/Downloads/lhs (1).csv", header = TRUE)
```

#Confidence Interval Calculation
```{r}
ci(lhs$wgtchg05, alpha = 0.05)
```

#Guided Questions
#1) The average weight change for the LHS participants is 3.71 kg. The value of the Standard Error is the 0.09.

#2) The degrees of freedom is 5468
```{r}
qt(0.975, df =5469-1)
```

#3) The lower 95 CI is 3.53 kg and the upper 95 CI is 3.89 kg.

#-------------------------------------------------

#Reclassifying Variables

#Reclassify numerical to categorical
```{r}
lhs$lung.fxn<-cut(lhs$FEVFVC3,
breaks=c(0, 69.99999999, 100),
labels=c("abnormal","normal"))
```

#Reclassify categorical to numerical
```{r}
lhs$marital0 <- as.factor(lhs$marital0)
str(lhs$marital0)
```


#Guided Questions
#1) Quotes are placed around the labels to signify how we want to label values from 0-69.9 and 70-100

#2) The name of the variable is martial0. It has the same name for both factor and interger form because we kept the name but redefined the data type.
