---
title: "Lesson 4 Lab"
author: "Patrick Torralba"
date: "2025-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

#Hypothesis Testing

##import data
```{r}
lhs <- read.csv(file =file.choose(), header = TRUE)
```


#two-sided t-test
```{r}
t.test(lhs$wgtchg05, conf.level=0.95, mu=0, alternative="two.sided")
```


#t-statistic value <--- this is our normal value of what we want to compare our t-statistic to
```{r}
qt(0.975, df = 5468)
```


##Guided Questions

#1) The sample mean weight change is 3.71 kg

#2) The test statistic is a t-statistic of 41.068 and the degrees of freedom is 5468.

#3) The resulting p-valye is 2.2e-16


---------------------------------------------

#Transforming a Numerical Variable in R

#Skewed Variable | This is a right-skewed graph
```{r}
hist(lhs$AV5CIGS, main="Histogram of average cigarettes at year 5",
xlab="average cigarettes per day")
```


###Note: A QQ plot is created to generate a normal distribution from the data to compare what the data values would look like if the data was normally distributed

## QQ Plot & Reference line
```{r}
qqnorm(lhs$AV5CIGS)
qqline(lhs$AV5CIGS)
```

##Log transform this variable 
```{r}
lhs$logAvgCigs5<-log(lhs$AV5CIGS)
```

## Guided Questions
#1)The first value is 2.30. There is missing data for rows 2,3, and 4.

```{r}
head(lhs)
```

#2) QQplot of log variable | No you cannot create a QQ plot of the new variable because these values go to infinity. This does not look symmetrical either.
```{r}
qnorm(lhs$logAvgCigs5)

```

#3 QQ plot of age variable | The age does not look symmetric
```{r}
log_age<-log(lhs$age)

hist(log_age)
```

