---
title: "Poisson Regression"
author: "Patrick Torralba"
format: 
  html:
    toc: true
    smooth-scroll: true
    number-sections: true
editor: visual
---

Set-up

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true

library(tidyverse)
library(flextable) ## for table extraction
library(gtsummary) ## for making tables
library(gt) ## for making tables
library(psych) ## calculate descriptive statistics
library(knitr) ## for making tables easily from
library(janitor) ## clean tables


```

```{r}
#| label: load data
#| message: false
#| warning: false
#| echo: false

lhs <- read.csv("lhs.csv")


fh <- read.csv("Framingham_dataset.csv")

```

# Section A. Count Data with Fixed Denominator (No Offset Term)

## Data Cleaning

```{r}
#| label: data-cln-lhs
#| warning: false
#| message: false
#| echo: true

lhs <- lhs %>% 
  mutate(cigs5 = if_else(AV5CIGS == 0.5, 1.0, AV5CIGS),
         Race = case_when(
           race == 1 ~ "White",
           race == 2 ~ "Black",
           race == 3 ~ "Asian",
           race == 4 ~ "Native American",
           race == 5 ~ "Other"
         ),
         Gender = if_else(
           AGENDER == "M", "Males", "Females"
         ),
        Marital = case_when(
          marital0 == 1 ~ "Never Married",
          marital0 == 2 ~ "Married",
          marital0 == 3 ~ "Widowed",
          marital0 == 4 ~ "Separated",
          marital0 == 5 ~ "Divorced"
        ),
        TreatmentGroup = fct_relevel(alphagroup, "UC")
  )
        


```

## cigs5 table

```{r}
#| label: tbl-cigs5
#| tbl-cap: "Cigs 5"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


lhs %>% 
  select(cigs5) %>% 
  filter(cigs5 == 0.5) %>% 
  tibble() %>% 
  flextable()



```

# Exploratory Data Analysis (EDA)

::: {.content-hidden when-format="html"}
Variables to pull for participant table:

-   race
-   f10cigs \[cigs per day at baseline\]
-   yeareduc \[years of education\]
-   age
-   bmi
-   WEIGHT
-   BLCIGAR
-   alphagroup \[randomized intervention groups\]
-   AGENDER
-   martial0
-   DRINKS BL
-   CO0 (expired carbon monoxide level at baseline)

I should make table 1 by treatment group. Make sure to show key demographics and baseline characteristics
:::

## Participant Demographic Table

```{r}
#| label: tbl-pt-demo
#| tbl-cap: "Participant Demographics and Baseline Characteristics"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


tbl_1 <- lhs %>% 
  select( Race, 
           f10cigs, 
           yeareduc, 
           age, 
           bmi, 
           alphagroup, 
           Gender, 
           Marital, 
           DRINKSBL,
          days2ded) %>% 
  tbl_summary(by = alphagroup,
              label = list(
                "f10cigs" = "Cigarettes Per Day at Baseline",
                "yeareduc" = "Years of Education",
                "age" = "Age",
                "bmi" = "BMI",
                "DRINKSBL" = "Alcoholic Drinks per Week at Baseline",
                "days2ded" = "Time (days) from baseline to death")
              ) %>% 
  add_overall() %>% 
  bold_labels()

tbl_1 ## I do not need to make it into a flex table because I am copy-pasting directly from the html to google docs


```

## cigs5 Histogram

```{r}
#| label: fig-cigs5-hist
#| fig-cap: "cigs5 Histogram"
#| fig-cap-location: top
#| warning: false
#| message: false
#| echo: true

ggplot(lhs, aes(x = cigs5)) +
  geom_histogram() + 
  labs(x = "Cigarettes Per Day at Annual Visit (Year 5)",
       y = "Count",
       title = "Distribution of Cigarettes Per Day at Annual Visit (Year 5)") +
  theme_minimal()


```

## cigs5 Descriptive Statistics

```{r}
#| label: tbl-cigs5-desc-stats
#| tbl-cap: "cigs5 Descriptive Statistics"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


describe(lhs$cigs5)


describe(lhs$cigs5,
         IQR = TRUE) %>% 
  as.data.frame() %>% 
  select(n, mean, median, IQR, sd, min, max, skew, kurtosis) %>%
  flextable() %>% 
  colformat_double(digits = 2) %>% 
  theme_booktabs() %>% 
  autofit() %>% 
  bold(part = "header") %>% 
  set_header_labels(            
    Variable = "Variable",
    n = "N",
    mean = "Mean",
    median = "Median",
    sd = "Std. Dev.",
    min = "Min",
    max = "Max",
    skew = "Skew",
    kurtosis = "Kurtosis"
  )


```

# Model 1: Effect of Treatment Group

## Poisson Regression Model

```{r}
#| label: tbl-trt-group-vs-cigs5
#| tbl-cap: "Poission Regression Model: The Effect of Treatment Group on Cigarette Smoking at 5 Years"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


trt_group_cigs5_model <- glm(cigs5 ~ TreatmentGroup, 
                             data = lhs,
                             family = poisson(link = "log"))

trt_group_cigs5_model 

trt_group_cigs5_model %>% ## this is the exponentiated form
  tbl_regression(
    exponentiate = TRUE,
    label = list(TreatmentGroup ~ "Treatment Group")
  )


trt_group_cigs5_model %>% ## this is the expected log form
  tbl_regression(
    exponentiate = FALSE,
    label = list(TreatmentGroup ~ "Treatment Group")
  )

```

The effect of the SI-P smoking intervention group is 0.67 relative to the usual group at year 5. In other words, the estimated Incidence Rate Ratio (IRR) is 0.67, meaning that the SI-P group smoked at a rate 33% lower than the usual care group. This finding is statistically significant given that our p-value is lower than alpha and the 95% confidence interval does not include 1.0 (p-value \< 0.001, CI: 0.66 - 0.68).

## Model Evaluation for Overdispersion Test

To evaluate the poisson regression, we need to review the ratio of the **residual deviance** to its **degree of freedoms**. **If the model ratio is super super greater than 1 then that is a sign of overdispersion.**

```{r}
#| label: overdispersion-test


model_ratio <- (trt_group_cigs5_model$deviance
/ trt_group_cigs5_model$df.residual)

model_ratio

model_ratio > 10

```

# Model 2: Effect of Treatment Group, Allowing for Overdispersion

How to deal with overdispersion? - 2 ways \* Quasi-Poisson Regression \* Negative Binomial Regression

## Quasi-Poisson Regression Model for Overdispersed Data

Use the Quasi-Possion Regression model to handle overdispersion

```{r}
#| label: tbl-quasi-possion-regression-model
#| tbl-cap: "Quasi-Poisson Regression Model"
#| tbl-cap-location: top


model1.qp <- glm(cigs5 ~ TreatmentGroup,
                 data = lhs,
                 family = quasipoisson(link = "log"))

model1.qp

model1.qp %>%  ## exponentiated form
  tbl_regression(
    exponentiate = TRUE,
    label = (TreatmentGroup ~ "Treatment Group")
  )

model1.qp %>%  ## expected log form
    tbl_regression(
    exponentiate = FALSE,
    label = (TreatmentGroup ~ "Treatment Group")
  )

```

# Section C. Binary Data Treated as Rates (With an Offset Term)

```{r}
#| label: data-cln-fh


fh <- fh %>% 
  mutate(followupYR = followup / 365.25,
         CHD = as.factor(if_else(chdfate == 1, "Developed CHD", "No CHD")),
         Sex = as.factor(if_else(sex == 1, "Male", "Female")))
         


```

## Exploratory Data Analysis (EDA)

I need to have the following information:
 * Cases of CHD
 * Follow-up time in person-years
 * CHD Incidence Rate per 100 person-years


```{r}
#| label: tbl-fh-IR
#| tbl-cap: "Incidence Rate Table for Framingham Dataset"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


fh_ir_tbl <- fh %>% 
  select(CHD, Sex, followupYR) %>% 
  group_by(Sex) %>% 
  summarize(
    CHD_cases = sum(CHD == "Developed CHD"),
    ft_py = sum(followupYR),
    CHD_IR = ((CHD_cases / ft_py)*100) ## CH Incidence rate calculation
            )  %>% 
  gt() %>% ## convert into gt() table
  tab_header(
      title = "Coronary Heart Disease Incidence Metrics") %>% 
  cols_label(
    CHD_cases = "Total Cases",
    ft_py = "Person-Years",
    CHD_IR = "Incidence Rate per 100 person-years"
            ) %>% 
  fmt_number( ## format Incidence Ratio
    columns = c(CHD_IR),
    decimals = 2
            ) %>% 
  fmt_number( ## format person-years
    columns = ft_py,
    decimals = 0
            ) 
  

fh_ir_tbl


```

