---
title: "Poisson Regression"
author: "Patrick Torralba"
format: 
  html:
    toc: true
    smooth-scroll: true
    number-sections: true
editor: visual
---

Set-up
```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true

library(tidyverse)
library(flextable) ## for table extraction
library(gtsummary) ## for making tables

```

```{r}
#| label: load data
#| message: false
#| warning: false
#| echo: false

lhs <- read.csv("lhs.csv")


```


# Section A. Count Data with Fixed Denominator (No Offset Term)

## Data Cleaning
```{r}
#| label: data-cln
#| warning: false
#| message: false
#| echo: true

lhs <- lhs %>% 
  mutate(cigs5 = if_else(AV5CIGS == 0.5, 1.0, AV5CIGS),
         Race = case_when(
           race == 1 ~ "White",
           race == 2 ~ "Black",
           race == 3 ~ "Asian",
           race == 4 ~ "Native American",
           race == 5 ~ "Other"
         ),
         Gender = if_else(
           AGENDER == "M", "Males", "Females"
         ),
        Marital = case_when(
          marital0 == 1 ~ "Never Married",
          marital0 == 2 ~ "Married",
          marital0 == 3 ~ "Widowed",
          marital0 == 4 ~ "Separated",
          marital0 == 5 ~ "Divorced"
         ),
  )
        


```

## cigs5 table
```{r}
#| label: tbl-cigs5
#| tbl-cap: "Cigs 5"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


lhs %>% 
  select(cigs5) %>% 
  filter(cigs5 == 0.5) %>% 
  tibble() %>% 
  flextable()



```

# Exploratory Data Analysis (EDA)

::: {.content-hidden when-format="html"}

Variables to pull for participant table:

* race
* f10cigs [cigs per day at baseline]
* yeareduc [years of education]
* age
* bmi
* WEIGHT
* BLCIGAR
* alphagroup [randomized intervention groups]
* AGENDER
* martial0
* DRINKS BL
* CO0 (expired carbon monoxide level at baseline)

I should make table 1 by treatment group. Make sure to show key demographics and baseline characteristics

:::

## Participant Demographic Table
```{r}
#| label: tbl-pt-demo
#| tbl-cap: "Participant Demographics and Baseline Characteristics"
#| tbl-cap-location: top
#| warning: false
#| message: false
#| echo: true


tbl_1 <- lhs %>% 
  select( Race, 
           f10cigs, 
           yeareduc, 
           age, 
           bmi, 
           alphagroup, 
           Gender, 
           Marital, 
           DRINKSBL,
          days2ded) %>% 
  tbl_summary(by = alphagroup,
              label = list(
                "f10cigs" = "Cigarettes Per Day at Baseline",
                "yeareduc" = "Years of Education",
                "age" = "Age",
                "bmi" = "BMI",
                "DRINKSBL" = "Alcoholic Drinks per Week at Baseline",
                "days2ded" = "Time (days) from baseline to death")
              ) %>% 
  add_overall() %>% 
  bold_labels()

tbl_1 ## I do not need to make it into a flex table because I am copy-pasting directly from the html to google docs


```

