---
title: "6451IP-Week-7-Problem-Set2"
author: "Patrick Torralba"
format: html
editor: visual
---

# Part 3: Carrying Out and Interpreting Analyses

```{r load data}

f6 <- read.csv("first6.csv")
  
```

```{r library}
library(tidyverse)
library(survival) #survival analysis
library(ggfortify) #survival analysis
library(gtsummary) #table 1
library(flextable) #save to docx
```

## Exploratory Data Analysis

### Data Cleaning

```{r data cleaning}
f6 <- f6 %>% 
    mutate(gender = as.factor(if_else(GENDER == 1, "Male", "Female", missing = NULL)),
         race = as.factor(case_when(RACE == 1 ~ "Latino",
                                    RACE == 2 ~ "African American",
                                    RACE == 3 ~ "White",
                                    RACE == 4 ~ "Other")),
         two_class = as.factor(if_else(twoclass == 1, "Two Class", "Three Class", missing = NULL)))
```

### Table 1

```{r q1}
table_1 <- f6 %>% 
  tbl_summary(by = two_class,
              include = c(AGE, CD4BL, gender, race, RNABL, two_class),
              label = list(AGE = "Age",
                           CD4BL = "Baseline CD4+",
                           gender = "Gender",
                           race = "Race",
                           RNABL = "Baseline Viral Overload")) %>% 
  modify_caption("**Table 1. Patient Characteristics**") %>% 
  add_overall() %>% 
  bold_labels()

table_1
```

```{r table 1 save as docx, eval = FALSE}
table_1 %>% 
  as_flex_table() %>% 
  save_as_docx(path = "Export/table_1.docx")

```

-   Note: need to add as_flex_table() function to convert from gt_table to save as document
    -   Otherwise, the document will appear empty.

### Kaplan-Meier plot: t2primary and primary vs twoclass

```{r q3 survival calculation}

q3 <- survfit(Surv(t2primary, primary) ~ two_class, data = f6)

```

-   Note: Surv() uses the form y \~ x.
    -   y = time to event outcome \<-- outcome variable
    -   y1 = time variable
    -   y2 = event indicator
-   x = twoclass \<-- explanatory variable
    -   Super important that you **calculate the survival first**
-   According to the data document, t2primary = time of first censoring. And, primary = Death occurs

```{r q3 graph}

autoplot(q3, conf.int = FALSE, censor.shape = "X") +
  labs(x  = "Time (in months)",
       y = "Survival Percentage",
       title = "Survival Analysis: Two-class vs Three-class",
       color = "Exposure")

```

### Formal Testing and Analysis

```{r q5 - use log_rank_test to determine significance between the two classes}

survdiff(Surv(t2primary, primary) ~ two_class, data = f6, rho = 0)


```

-   Note: To determine if there's statistical significance between two groups in a survival analysis, you must do the survival calculation **and use survdiff() and set rho = 1 or rho = 0**
-   rho = 0 \<-- log-rank test
    -   This test weights all points equally
-   rho = 1 \<-- Wilcoxon test
    -   This test puts more weight on early events
    -   both have the same null hyp. Null hyp: survival identical in both groups.
